---
knitr:
  opts_chunk:
    cache.path: "../_cache/cls-ensembles/"
---

# Classification Ensembles {#sec-cls-ensembles}


```{r}
#| label: cls-ensembles-setup
#| include: false
source("../R/_common.R")
source("../R/_themes.R")
source("../R/_themes_ggplot.R")
source("../R/_themes_gt.R")

# ------------------------------------------------------------------------------
# Required packages

library(tidymodels)
library(mirai)
library(spatialsample)
library(future)
library(patchwork)
library(gt)
library(gtExtras)
library(lorax)
library(bonsai)
library(probably)
library(tidyposterior)
library(rules)
library(paletteer)
library(baguette)

# ------------------------------------------------------------------------------
# set options

tidymodels_prefer()
set_options()
daemons(parallel::detectCores())
theme_set(thm_lt)

source("../R/compare_models.R")

# ------------------------------------------------------------------------------
# load data

load("../RData/forested_data.RData")

# ------------------------------------------------------------------------------

cls_mtr <- metric_set(brier_class, roc_auc, pr_auc, mn_log_loss)
```

## Bagging {#sec-cls-bag}


```{r}
#| label: forest-bag-cart
#| include: false
#| cache: true

cart_bag_spec <-
  bag_tree(mode = "classification") |>
  set_engine("rpart", times = 50)

set.seed(285)
forest_cart_bag_res <-
  cart_bag_spec |>
  fit_resamples(
    class ~ .,
    resamples = forested_rs,
    control = control_resamples(
      save_pred = TRUE,
      save_workflow = TRUE
    ),
    metrics = cls_mtr
  )

forest_cart_bag_best <- select_best(forest_cart_bag_res, metric = "brier_class")

forest_cart_bag_pred <-
  forest_cart_bag_res |>
  collect_predictions(parameters = forest_cart_bag_best) |>
  mutate(wflow_id = "Bagged CART") |>
  relocate(wflow_id)

forest_cart_bag_mtr_ex <-
  forest_cart_bag_pred |>
  cls_mtr(class, estimate = .pred_class, .pred_Yes)
```


## Random Forest {#sec-cls-rand-forest}

```{r}
#| label: forest-rf-ranger
#| include: false
#| cache: true

rngr_spec <-
  rand_forest(
    mode = "classification",
    mtry = tune(),
    trees = 500,
    min_n = tune()
  )

set.seed(526)
forest_rngr_res <-
  rngr_spec |>
  tune_grid(
    class ~ .,
    resamples = forested_rs,
    grid = 25,
    control = control_grid(
      save_pred = TRUE,
      save_workflow = TRUE
    ),
    metrics = cls_mtr
  )

forest_rngr_best <- select_best(forest_rngr_res, metric = "brier_class")

forest_rngr_pred <-
  forest_rngr_res |>
  collect_predictions(parameters = forest_rngr_best) |>
  mutate(wflow_id = "Random Forest") |>
  relocate(wflow_id)

forest_rngr_mtr_ex <-
  forest_rngr_pred |>
  cls_mtr(class, estimate = .pred_class, .pred_Yes)
```


```{r}
#| label: forest-rf-obl
#| include: false
#| cache: true

rf_obl_spec <-
  rand_forest(
    mode = "classification",
    mtry = tune(),
    trees = 500,
    min_n = tune()
  ) |>
  set_engine("aorsf")

set.seed(545)
forest_rf_obl_res <-
  rf_obl_spec |>
  tune_grid(
    class ~ .,
    resamples = forested_rs,
    grid = 25,
    control = control_grid(
      save_pred = TRUE,
      save_workflow = TRUE
    ),
    metrics = cls_mtr
  )

forest_rf_obl_best <- select_best(forest_rf_obl_res, metric = "brier_class")

forest_rf_obl_pred <-
  forest_rf_obl_res |>
  collect_predictions(parameters = forest_rf_obl_best) |>
  mutate(wflow_id = "Oblique Random Forest") |>
  relocate(wflow_id)

forest_rf_obl_mtr_ex <-
  forest_rf_obl_pred |>
  cls_mtr(class, estimate = .pred_class, .pred_Yes)
```

## Boosting {#sec-cls-boosting}


```{r}
#| label: forest-boost-c5
#| include: false
#| cache: true

c5_spec <-
  boost_tree(
    mode = "classification",
    trees = tune(),
    min_n = tune()
  ) |>
  set_engine("C5.0")

set.seed(498)
forest_c5_res <-
  c5_spec |>
  tune_grid(
    class ~ .,
    resamples = forested_rs,
    grid = 25,
    control = control_grid(
      save_pred = TRUE,
      save_workflow = TRUE
    ),
    metrics = cls_mtr
  )

forest_c5_best <- select_best(forest_c5_res, metric = "brier_class")

forest_c5_pred <-
  forest_c5_res |>
  collect_predictions(parameters = forest_c5_best) |>
  mutate(wflow_id = "Boosted C5.0") |>
  relocate(wflow_id)

forest_c5_mtr_ex <-
  forest_c5_pred |>
  cls_mtr(class, estimate = .pred_class, .pred_Yes)
```

```{r}
#| label: forest-boost-xgb
#| include: false
#| cache: true

xgb_spec <-
  boost_tree(
    mode = "classification",
    trees = 1000,
    mtry = tune(),
    min_n = tune(),
    tree_depth = tune(),
    loss_reduction = tune(),
    learn_rate = tune(),
    stop_iter = 5
  )

set.seed(754)
forest_xgb_res <-
  xgb_spec |>
  tune_grid(
    class ~ .,
    resamples = forested_rs,
    grid = 25,
    control = control_grid(
      save_pred = TRUE,
      save_workflow = TRUE
    ),
    metrics = cls_mtr
  )

forest_xgb_best <- select_best(forest_xgb_res, metric = "brier_class")

forest_xgb_pred <-
  forest_xgb_res |>
  collect_predictions(parameters = forest_xgb_best) |>
  mutate(wflow_id = "XGBoost") |>
  relocate(wflow_id)

forest_xgb_mtr_ex <-
  forest_xgb_pred |>
  cls_mtr(class, estimate = .pred_class, .pred_Yes)
```

```{r}
#| label: forest-boost-lgb
#| include: false
#| cache: true

lgb_spec <-
  boost_tree(
    mode = "classification",
    trees = 1000,
    mtry = tune(),
    min_n = tune(),
    tree_depth = tune(),
    loss_reduction = tune(),
    learn_rate = tune(),
    stop_iter = 5
  ) |>
  set_engine("lightgbm")

set.seed(931)
forest_lgb_res <-
  lgb_spec |>
  tune_grid(
    class ~ .,
    resamples = forested_rs,
    grid = 25,
    control = control_grid(
      save_pred = TRUE,
      save_workflow = TRUE
    ),
    metrics = cls_mtr
  )

forest_lgb_best <- select_best(forest_lgb_res, metric = "brier_class")

forest_lgb_pred <-
  forest_lgb_res |>
  collect_predictions(parameters = forest_lgb_best) |>
  mutate(wflow_id = "LightGBM") |>
  relocate(wflow_id)

forest_lgb_mtr_ex <-
  forest_lgb_pred |>
  cls_mtr(class, estimate = .pred_class, .pred_Yes)
```

```{r}
#| label: forest-boost-ctb
#| include: false

load("../RData/forest_catboost.RData")
```

### Classical Boosting {#sec-cls-classic-boosting}

### Modern Boosting {#sec-cls-modern-boosting}

## Bayesian Additive Regression Trees {#sec-cls-bart}

```{r}
#| label: forest-bart
#| include: false
#| cache: true

bart_spec <-
  parsnip::bart(
    mode = "classification",
    trees = 500,
    prior_terminal_node_coef = tune(),
    prior_terminal_node_expo = tune()
  ) |>
  set_engine("dbarts", keepevery = 10)

set.seed(637)
forest_bart_res <-
  bart_spec |>
  tune_grid(
    class ~ .,
    resamples = forested_rs,
    grid = 25,
    control = control_grid(
      save_pred = TRUE,
      save_workflow = TRUE
    ),
    metrics = cls_mtr
  )

forest_bart_best <- select_best(forest_bart_res, metric = "brier_class")

forest_bart_pred <-
  forest_bart_res |>
  collect_predictions(parameters = forest_bart_best) |>
  mutate(wflow_id = "BART") |>
  relocate(wflow_id)

forest_bart_mtr_ex <-
  forest_bart_pred |>
  cls_mtr(class, estimate = .pred_class, .pred_Yes)
```


## Rule-Based Ensembles {#sec-cls-rule-ensembles}

### C5.0 Rules {#sec-cls-c50-rules}

### RuleFit {#sec-cls-rule-fit}

## Comparisons to Previous Models {#sec-cls-ensemble-compare}

```{r}
#| label: cls-ensemble-compare
#| include: false
load("../RData/forest_logistic_set_res.Rdata")
load("../RData/forest_glmnet_set_res.Rdata")
load("../RData/forest_linear_res.Rdata")

load("../RData/forest_fda_set_res.Rdata")
load("../RData/forest_knn_set_res.Rdata")

load("../RData/forest_mlp_set_res.Rdata")
load("../RData/forest_autoint_set_res.Rdata")
load("../RData/forest_saint_set_res.Rdata")
load("../RData/forest_tabpfn_set_res.Rdata")
load("../RData/forest_tabicl_set_res.Rdata")

load("../RData/forest_cart_set_res.RData")
load("../RData/forest_c5_set_res.RData")
load("../RData/forest_cit_set_res.RData")
load("../RData/forested_obl_set_res.RData")
load("../RData/forest_c5rules_set_res.RData")


forest_ensemble_set_res <-
  as_workflow_set(
    Bagged_CART = forest_cart_bag_res |> butcher::butcher(),
    RF = forest_rngr_res |> butcher::butcher(),
    Oblique_RF = forest_rf_obl_res |> butcher::butcher(),
    Boost_C5.0 = forest_c5_res |> butcher::butcher(),
    XGBoost = forest_xgb_res |> butcher::butcher(),
    LightGBM = forest_lgb_res |> butcher::butcher(),
    BART = forest_bart_res |> butcher::butcher()
  )

all_sets <-
  dplyr::bind_rows(
    forest_fda_set_res,
    forest_knn_set_res,
    forest_glmnet_set_res,
    forest_logistic_set_res,
    forest_mlp_set_res,
    forest_tabpfn_set_res,
    forest_tabicl_set_res,
    forest_autoint_set_res,
    forest_saint_set_res,
    forested_obl_set_res,
    forest_cart_set_res,
    forest_c5_set_res,
    forest_c5rules_set_res,
    forest_cit_set_res,
    forest_ensemble_set_res,
    forest_ctb_set_res
  )
```

```{r}
#| label: cls-ensemble-compare-bayes
#| include: false
#| cache: false
set.seed(905)
ensemble_compare <- compare_models(
  all_sets,
  chains = 10,
  cores = 10,
  iter = 5000
)
ensemble_compare

ensemble_best <- ensemble_compare$model[is.na(ensemble_compare$probability)]

ensemble_subset <-
  ensemble_compare |>
  slice_min(median, by = c(class, engine)) |>
  mutate(
    model = as.character(model),
    model = ifelse(grepl("basic_", model), "Oblique Tree", model),
    model = ifelse(
      grepl("logistic_mle", model),
      "Logistic (unpenalized)",
      model
    ),
    model = ifelse(grepl("1L", model), "MLP (1 Layer)", model),
    model = ifelse(grepl("2L", model), "MLP (2 Layers)", model),
    model = gsub("glmnet", "Logistic (penalized)", model),
    model = ifelse(grepl("basic_", model), "Oblique Tree", model),
    model = ifelse(grepl("tab_pfn", model), "TabPFN", model),
    model = ifelse(grepl("autoint", model), "AutoInt", model),
    model = gsub("_", " ", model),
    model = factor(model),
    model = reorder(model, median)
  )

ensemble_best_brier <- round(
  ensemble_subset$median[ensemble_subset$model == ensemble_best],
  3
)
```

How well do these models perform relative to previously discussed techniques?  @fig-cls-ensemble-compare visualizes the Brier score results across models. The Bayesian methods in @sec-compare-resample-bayes were used to estimate the probability distributions for each of the best candidates for each model. The intervals show the 0.05, 0.50, and 0.95 quantiles. Currently, the `r ensemble_best` model has the smallest Brier score with a Bayesian point estimate of `r ensemble_best_brier`.

```{r}
#| label: fig-cls-ensemble-compare
#| echo: false
#| warning: false
#| out-width: 70%
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Comparisons of current forestation models. The intervals are 90% credible intervals. The bars on the right show the probability of practical equivalence to the current best model, assuming that a difference in Brier scores of +/- 0.01 is practical."
#| fig-alt: "Two-panel figure comparing all forestation models discussed so far. The left panel is a dot-and-whisker plot of Brier scores with 90% credible intervals, ordered from best to worst. Tree-based models cluster at higher Brier scores than the leading models. The right panel shows horizontal bars for the probability of practical equivalence, indicating that single tree models have virtually no chance of matching the currently best-performing models."
p_brier <-
  ensemble_subset |>
  ggplot(aes(median, model)) +
  geom_point() +
  geom_errorbar(aes(xmin = lower, xmax = upper), width = 1 / 2) +
  labs(y = NULL, x = "Brier Score") +
  theme(
    plot.margin = margin(t = 10, r = 2, b = 10, l = 10, unit = "pt")
  )

p_brier_diff <-
  ensemble_subset |>
  ggplot(aes(pract_equiv, model)) +
  geom_bar(stat = "identity", aes(fill = pract_equiv), show.legend = FALSE) +
  labs(y = NULL, x = "Pr[Equiv]") +
  scale_x_continuous(breaks = round((0:2) / 2, 2), limits = 0:1) +
  scale_fill_viridis_c(direction = -1) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 2, unit = "pt")
  )

p_brier + p_brier_diff + plot_layout(widths = c(4, 1))
```


## Chapter References {.unnumbered}


```{r}
#| label: cls-ensemble-results
#| include: false
save(forest_ensemble_set_res, file = "../RData/forest_ensemble_set_res.RData")
```
