---
knitr:
  opts_chunk:
    cache.path: "../_cache/cls-neural-nets/"
---

# Neural Network Classifiers {#sec-cls-neural-nets}

```{r}
#| label: cls-neural-nets-setup
#| include: false

source("../R/_common.R")

# ------------------------------------------------------------------------------

library(tidymodels)
library(torch)
library(mirai)
library(spatialsample)
library(embed)
library(bestNormalize)
library(gt)
library(probably)
library(patchwork)
library(paletteer)
library(brulee)
library(tidyposterior)

# ------------------------------------------------------------------------------
# set options

tidymodels_prefer()
theme_set(theme_bw())
set_options()
theme_set(theme_bw())
daemons(parallel::detectCores())

source("../R/compare_models.R")

cls_mtr <- metric_set(brier_class, roc_auc, pr_auc, mn_log_loss)

# ------------------------------------------------------------------------------
# load data

load("../RData/forested_data.RData")
load("../RData/optimizers.RData")
load("../RData/dat_2d.RData")

# ------------------------------------------------------------------------------
# Gradient descent helpers

source("../R/setup_2D_GD.R")

# ------------------------------------------------------------------------------

aligned <- function(x, ...) {
  n <- nrow(x)
  p <- ncol(x)

  fmt <- format(x, ...)
  fmta <- format(abs(x), ...)
  nonneg <- x < 0
  signs <- ifelse(nonneg, "-", "+")
  vars <- colnames(x)

  amps <- rep(" & ", n)

  just <- "rr"
  res <- cbind(rownames(x), amps)
  res <- cbind(res, fmt[, 1], amps)
  for (i in 2:p) {
    symb <- rep(vars[i], n)
    vals <- gsub("^ +", "+", fmt[, i])
    res <- cbind(res, vals, amps, symb)
    just <- c(just, "rl")
    if (i < p) {
      res <- cbind(res, amps)
    }
  }
  res <- cbind(res, c(rep("\\\\", n - 1), ""))
  justs <- paste(just, collapse = "")
  colnames(res) <- rownames(res) <- NULL

  cat(paste0("\n\\begin{array}{", justs, "}\n"))
  tt <- apply(res, 1, function(x) cat(x, "\n"))
  cat(paste0("\\end{array}\n"))
}
```

This chapter contains _three_ sections on neural networks. The first describes the most basic version of these models, called the _multilayer perceptron_. The fundamentals of these models are discussed in this section. The second section explores various methodologies to enhance the performance of neural network models for tabular data, similar to their effectiveness in predicting images and text. At the time of this writing, substantial research on this topic is underway in the field. This section will also discuss one of the most effective tools for most deep learning models: attention-based transformers. After this, we'll consider _tabular foundational models_, which are very effective pretrained neural networks. 

## Multilayer Perceptrons {#sec-cls-nnet}

Neural networks are complex models comprised of multiple hierarchies of _latent variables_ designed to represent different motifs in the data. The model's architecture consists of a series of _layers_, each containing a set of units. Neural networks begin with the input layer that contains units representing our predictors. After this are usually one or more _hidden layers_, each containing a preset number of hidden units. These are artificial features that are influenced by the layer before.  For example, if there were four predictors and the first hidden layer had two hidden units, each hidden unit is defined by an equation that includes all four predictors. In the broadest sense, it is similar to how, in PCA, each principal component is a linear combination of the predictors. However, the similarities between neural networks and PCA end there. 

These connections between units on different layers continue as layers are added. Eventually, the last layer is the _output layer_. For classification, this usually has as many units as there are class levels in the outcome. @fig-nnet-cls shows a general architectural diagram for a basic model with a single hidden layer, where the lines indicate that a model parameter connects the nodes. 


```{r}
#| label: fig-nnet-cls
#| echo: false
#| out-width: 40%
#| fig-width: 4
#| fig-height: 5
#| fig-cap: Architecture of a simple neural network with a single layer of hidden units. Lines represent slope parameters (a.k.a., weights) between variables. Architectures typically contain intercept parameters (a.h.a. biases), but these are not shown in this diagram. 
#| fig-alt: "Diagram of a multilayer perceptron with an input layer of predictor nodes on the left, a single hidden layer of nodes in the middle, and an output layer with nodes for each class on the right. Lines connect every node in each layer to every node in the next layer, representing model parameters."

knitr::include_graphics("../premade/mlp.svg")
```

Traditionally, the connections that come into a hidden unit are combined using a linear predictor: each incoming unit has a corresponding slope, called a "weight" in the neural network literature^[For consistency, we'll use "slopes" and "intercepts" here.]. The linear combinations commonly include an intercept term (called a "bias"). These linear predictors are embedded within a nonlinear "activation" function ($\sigma$) to enable complex patterns. An example of an activation function would be a sigmoidal function, much like the logistic function used in @eq-logistic. 

::: {.note-box}
In this section, we will primarily be concerned with multilayer perceptrons (MLPs), a subset of the broader class of neural networks. MLPs are feed-forward models, meaning each layer only connects to the subsequent layer; there are no feedback loops where future units connect to units in previous layers.

As a counterexample, residual networks [@he2016deep] are connected to previous layers (a.k.a. skip layers); therefore, they are not multilayer perceptrons. Another popular example is the long-term short memory model for sequential data such as time series [@hochreiter1997long].
:::

<br> 

Let's walk through the network structure, starting at the input layer. We have numeric predictors $x_1, \ldots x_p$. These will be mathematically connected to the next layer containing $m$ hidden units $H_\ell$ ($\ell=1, \ldots m$). First, we create a separate linear predictor for hidden unit $H_l$ denoted as $\beta_{\ell 0} + \beta_{\ell 1} x_{1} + \ldots + \beta_{\ell p} x_{p}$. We then embed this inside the activation function $\sigma()$ used for this layer so that there is a nonlinear relationship between the predictors and their new representation. Visually^[
This representation and the one below were adapted from Izaak Neutelings's excellent work, which can be found at [`https://tikz.net/neural_networks/`](https://tikz.net/neural_networks/) and is licensed under a CC BY-SA 4.0.]:

```{r}
#| label: nnet-start
#| echo: false
#| out-width: 80%
#| fig-width: 4
#| fig-height: 5
#| fig-alt: "Diagram showing the input-to-hidden-layer portion of a neural network. Predictor nodes on the left connect to a single hidden unit on the right using thicker lines. The linear combination of inputs is passed through an activation function to produce the hidden unit's output."

knitr::include_graphics("../premade/nnet_start.svg")
```

As we add more layers, each of the first set of hidden units is similarly connected to the hidden units in the next layer, and so on. The layers do not have to use the same activation function. Also, the layers can be specialized, such as the image processing layers shown in @fig-vgg16. For each hidden layer, we have to specify the number of units and the activation function. These, along with the number of layers, are tuning parameters. 

Note that the units in the subsequent hidden layer are nested inside linear combinations of the previous hidden layers (plus the input units). These models comprise sets of repeatedly nested nonlinear functions and can be very complex yet remain mathematically tractable. 

::: {.important-box}
The hidden units ($H_{j}$) should be considered _embeddings_, previously described in @sec-embeddings. In this chapter, the term is ubiquitous, and we should take a minute to revisit it. We can think of an embedding as a transformation of an existing variable or data vector into a new representation (which can have a different dimensionality). That definition is very broad, and the term is most frequently used in the neural network literature. Some embeddings are fixed lookups (e.g., dummy variables), some are functions of a single input (splines), and some are functions of the entire input set (PCA features). More complex networks go far beyond basic hidden layers and units, and there will be additional terms that we'll call embeddings. 
::: 

For the final layer, the previous set of hidden units is often connected to the _output units_ using a linear activation. For classification, there are often $C$ output units (i.e., one for each class):

```{r}
#| label: nnet-end
#| echo: false
#| out-width: 70%
#| fig-width: 4
#| fig-height: 5
#| fig-alt: "Diagram showing the hidden-to-output-layer portion of a neural network. Hidden unit nodes on the left connect via weighted lines to C output units on the right, one per class. A linear activation is used for the output layer, and the raw outputs are then normalized via the softmax function."

knitr::include_graphics("../premade/nnet_end.svg")
```

Since a linear activation is used in the final layer, the output units ($Y$) are unlikely to be in the proper units for classification. To remedy this, the raw outputs are normalized to be probability-like via the softmax function: 

$$
\pi_{k} = \frac{\exp(Y_{k})}{\sum\limits_{k=1}^C \exp(Y_{k})}
$$

In this section, we'll describe the important elements of multilayer perceptrons, such as the different types of activation functions. We'll walk through a small but specific model. 

In later subsections, we'll also spend considerable time discussing various methods for training MLPs. There are also discussions of modern deep learning models specifically designed for tabular data. These, along with powerful _foundational models_, are discussed below. 

### Activation Functions {#sec-nnet-activation}

```{r}
#| label: act-grid
#| include: false

lp_grid <- seq(-3, 3, length.out = 100)

act_functions <-
  bind_rows(
    tibble(
      input = lp_grid,
      output = binomial()$linkinv(lp_grid),
      activation = "logistic"
    ),
    tibble(input = lp_grid, output = tanh(lp_grid), activation = "tanh"),
    tibble(
      input = lp_grid,
      output = as.numeric(nnf_relu(lp_grid)),
      activation = "relu"
    ),
    tibble(
      input = lp_grid,
      output = as.numeric(nnf_elu(lp_grid)),
      activation = "elu"
    ),
    tibble(
      input = lp_grid,
      output = lp_grid - tanh(lp_grid),
      activation = "tanhshrink"
    ),
    tibble(
      input = lp_grid,
      output = as.numeric(nnf_softshrink(lp_grid)),
      activation = "softshrink"
    ),
    tibble(
      input = lp_grid,
      output = as.numeric(nnf_gelu(lp_grid)),
      activation = "gelu"
    ),
    tibble(
      input = lp_grid,
      output = as.numeric(nnf_silu(lp_grid)),
      activation = "silu"
    ),
    tibble(
      input = lp_grid,
      output = as.numeric(nn_softplus()(lp_grid)),
      activation = "softplus"
    )
  )
```

Historically, the most commonly used nonlinear activation functions were sigmoidal. This means that the linear combination of the elements in the previous layer was converted into a "squashed" format, typically bounded (e.g., logistic activation values range between 0 and 1). @fig-sigmoid-activations shows several such functions. The logistic (@eq-logistic) function and hyperbolic tangent are fairly similar in shape and were typical defaults. 

```{r}
#| label: fig-sigmoid-activations
#| echo: false
#| out-width: 65%
#| fig-width: 6
#| fig-height: 2
#| fig-cap: "Different sigmoidal activation functions."
#| fig-alt: "Faceted line plot of three sigmoidal activation functions: logistic, tanh, and tanhshrink. The logistic function is bounded between zero and one with an S-shape. The tanh function has a similar shape but ranges from -1 to 1. The tanhshrink function is near zero for small inputs and becomes approximately linear for large absolute inputs."

sig_lvl <- c("logistic", "tanh", "tanhshrink")

act_functions |>
  filter(grepl("(tan)|(sig)|(logistic)", activation)) |>
  mutate(activation = factor(activation, levels = sig_lvl)) |>
  ggplot(aes(input, output)) +
  geom_line() +
  facet_wrap(~activation, scale = "free_y", nrow = 1) +
  theme_bw()
```

One issue with bounded sigmoidal shapes, particularly the logistic function, is that they can become problematic when used in conjunction with many hidden layers. Since the logistic function is on [0, 1], repeated multiplication of such values can cause their values to converge to be very close to zero (as seen in @sec-naive-bayes). This is also true of their gradients, leading to the _vanishing gradient problem_. This is likely not an issue with shallow networks that have a handful of hidden layers.

The tanhshrink function diminishes the impact of inputs that are in a neighborhood around zero, where it is sigmoidal. For larger input values, the output is linear in the input (with an offset of $\pm 1$).

In modern neural networks, the rectified linear unit (ReLU) is the default activation function. As shown in @fig-relu-activations, the function resembles the hinge functions used by MARS: to the left of zero, the output is zero; otherwise, it is linear. Much like MARS, this isolates a region of the predictor space. However, there are a few key differences. MARS hinge functions affect a single predictor at a time, whereas neural network activation functions act on a linear combination of parameters. Also, neural networks with multiple layers recursively activate linear combinations inside the prior layers (and so on). 

```{r}
#| label: fig-relu-activations
#| echo: false
#| out-width: 90%
#| fig-width: 8
#| fig-height: 2
#| fig-cap: "Different ReLU-based activation functions."
#| fig-alt: "Faceted line plot of five ReLU-based activation functions: relu, softplus, gelu, silu, and elu. All are approximately zero or negative for negative inputs and approximately linear for positive inputs. ReLU has a sharp corner at zero, while softplus, gelu, silu, and elu provide smooth approximations with slightly different curvatures near the origin."

relu_lvl <- c("relu", "softplus", "gelu", "silu", "elu")

act_functions |>
  filter(grepl("(l[uU]$)|(softplus)", activation)) |>
  mutate(activation = factor(activation, levels = relu_lvl)) |>
  ggplot(aes(input, output)) +
  geom_hline(yintercept = 0, lty = 3) +
  geom_line() +
  facet_wrap(~activation, nrow = 1) +
  theme_bw()
```

There are various modified versions of ReLU that emulate the same form but have continuous derivatives. These are also shown in @fig-relu-activations, including softplus [@dugas2000incorporating;@glorot2011deep], gelu [@hendrycks2016gaussian], silu [@elfwing2018sigmoid],  elu [@clevert2015fast], and others. ReLU and its offspring activation functions appear to address the vanishing-gradient problem. 

```{r}
#| label: nnet-ex-comps
#| include: false
#| cache: true

x <- seq(-3, 3, length.out = 100)
grid <- crossing(A = x, B = x)

set.seed(123)
fit_1 <- brulee_mlp(
  class ~ A + B,
  data = dat_2d_train,
  hidden_units = 4,
  penalty = 0,
  learn_rate = 0.001,
  activation = "relu",
  optimizer = "ADAMw",
  batch_size = 16L,
)

coefs <- stats::coef(fit_1)

grid_pred <-
  fit_1 |>
  predict(grid, type = "prob") |>
  bind_cols(grid)

get_elements <- function(x, model) {
  xtbl <- x
  x <- as.matrix(x)
  lp_01 <- x %*% t(coefs$model.0.weight) + coefs$model.0.bias
  colnames(lp_01) <- paste0("lp_", 1:ncol(lp_01))
  layer_01 <- as.matrix(torch::nnf_relu(lp_01))
  colnames(layer_01) <- paste0("relu_", 1:ncol(layer_01))
  lp_12 <- layer_01 %*% t(coefs$model.2.weight) + coefs$model.2.bias
  colnames(lp_12) <- paste0("output_", 1:ncol(lp_12))
  prob <- exp(lp_12) / sum(exp(lp_12))
  bind_cols(xtbl, as_tibble(lp_01), as_tibble(layer_01), as_tibble(lp_12))
}

components <- NULL
for (i in seq_len(nrow(grid))) {
  components <- bind_rows(components, get_elements(grid[i, ], fit_1))
}
```

Let's use an example data set with a simple MLP to demonstrate how ReLU activation works. @fig-nnet-ex displays the results of a model with a single layer comprising four hidden units, using ReLU activation. The model converged in `r fit_1$best_epoch` epochs and does a good job discriminating between the two classes. 

```{r}
#| label: fig-nnet-ex
#| echo: false
#| out-width: 60%
#| fig-width: 5
#| fig-height: 4.4
#| warning: false
#| fig-cap: "An example data set with two predictors and two classes. The black line is the class boundary from an MLP model using four hidden units with ReLU activation."
#| fig-alt: "Scatter plot of a two-class data set with predictor A on the x-axis and predictor B on the y-axis. The two classes form partially overlapping clouds. A black contour line shows the decision boundary from an MLP with four ReLU hidden units, which traces a path similar to a piecewise-linear boundary separating the classes."

grid_pred |>
  ggplot(aes(A, B)) +
  geom_point(data = dat_2d_train, aes(col = class), alpha = 1 / 3) +
  geom_contour(aes(z = .pred_one), breaks = 0.5, col = "black") +
  coord_fixed(ratio = 1.2) +
  scale_color_manual(values = c("#007FFFFF", "#FF7F00FF"))
```

With two predictors and four hidden units, the first set of coefficients for the model is four sets of linear predictors (i.e., an intercept and two slopes) for each of the four hidden units. The predictors were normalized to have the same mean and variance (zero and one, respectively). The four equations produced during training were: 

```{r}
#| label: nnet-ex-first
#| echo: false
#| results: "asis"

align_equations <- function(
  coefs,
  symbols,
  outputs,
  digits = 2,
  ref = character(0)
) {
  n_eq <- nrow(coefs)
  n_terms <- ncol(coefs)
  n_align <- 1 + n_terms

  lines <- character(n_eq)

  for (i in seq_len(n_eq)) {
    parts <- character(n_terms + 1)
    parts[1] <- paste0(outputs[i], " &{}={}")

    for (j in seq_len(n_terms)) {
      val <- coefs[i, j]
      num_str <- formatC(abs(val), format = "f", digits = digits)
      sym_str <- if (nzchar(symbols[j])) paste0("\\,", symbols[j]) else ""

      if (j == 1) {
        sign_str <- if (val < 0) "-" else ""
        parts[j + 1] <- paste0("& ", sign_str, " & ", num_str, sym_str)
      } else {
        sign_str <- if (val < 0) "{}-{}" else "{}+{}"
        parts[j + 1] <- paste0("& ", sign_str, " & ", num_str, sym_str)
      }
    }

    lines[i] <- paste(parts, collapse = " ")
  }

  lines[-n_eq] <- paste0(lines[-n_eq], " \\\\")

  if (length(ref) > 0) {
    ending <- cli::format_inline("$$ {{#eq-{ref}}}\n")
  } else {
    ending <- "$$\n"
  }

  cat("$$\n")
  cat("\\begin{alignat*}{", n_align, "}\n", sep = "")
  cat(paste(lines, collapse = "\n"), "\n", sep = "")
  cat("\\end{alignat*}\n")
  cat(ending)
}

align_equations(
  cbind(coefs$model.0.bias, coefs$model.0.weight),
  c("", "A", "B"),
  paste0("H_{", 1:4, "}")
)
```

Based on the absolute magnitude of their slope coefficients, the second hidden unit appears to have the most dynamic effect on the results, followed by the fourth. The third hidden unit has a negligible function of the two predictors. $H_1$ and $H_4$ share similar patterns but differ in effect magnitude. The top of @fig-nnet-ex-layer-1 displays a heatmap illustrating how these four artificial features vary across the two-dimensional predictor space (before activation). The black line segments show the contour at zero. 

```{r}
#| label: fig-nnet-ex-layer-1
#| dev: "ragg_png"
#| echo: false
#| out-width: 80%
#| fig-width: 7
#| fig-height: 5
#| dev: "ragg_png"
#| warning: false
#| fig-cap: "Heatmaps of the four linear predictors and their ReLU values across the predictor space. The black line emphasizes the location where the linear predictor is zero."
#| fig-alt: "Two rows of four heatmaps each. The top row shows the raw linear predictor values for four hidden units across the A-B predictor space, each with a black contour line at zero dividing positive and negative regions. The bottom row shows the same values after ReLU activation, where negative regions are set to zero, and positive regions retain their values."

lp_layer <-
  components |>
  dplyr::select(-starts_with("relu_"), -starts_with("output")) |>
  pivot_longer(cols = c(starts_with("lp_"))) |>
  mutate(
    split_up = map(name, ~ strsplit(.x, split = "_")[[1]]),
    name = map_chr(split_up, ~ paste0("H[", .x[2], "]"))
  ) |> 
  ggplot(aes(A, B)) +
  geom_tile(aes(fill = value), show.legend = TRUE) +
  geom_contour(aes(z = value), breaks = 0, col = "black") +
  facet_wrap(~name, nrow = 1, labeller = label_parsed) +
  scale_fill_gradient2(high = "#5D74A5FF") +
  # coord_fixed() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  ) +
  labs(title = "Linear Predictor")

relu_layer <-
  components |>
  dplyr::select(-starts_with("lp"), -starts_with("output")) |>
  pivot_longer(cols = c(starts_with("relu_"))) |>
  mutate(
    split_up = map(name, ~ strsplit(.x, split = "_")[[1]]),
    name = map_chr(split_up, ~ paste0("H[", .x[2], "]"))
  ) |> 
  ggplot(aes(A, B)) +
  geom_tile(aes(fill = value), show.legend = TRUE) +
  geom_contour(aes(z = value), breaks = 0, col = "5AAE61FF") +
  facet_wrap(~name, nrow = 1, labeller = label_parsed) +
  scale_fill_gradient2(high = "#5D74A5FF") +
  # coord_fixed() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  ) +
  labs(title = "ReLU")

lp_layer / relu_layer
```

```{r}
#| label: nnet-ex-weights
#| include: false
weight_sum <-
  cbind(coefs$model.0.bias, coefs$model.0.weight)[, -1] |>
  abs() |>
  apply(1, sum)

top_weights <- sort(seq_along(weight_sum)[order(weight_sum, decreasing = TRUE)][
  1:2
])
top_weights <- sort(paste0("$H_{", top_weights, "}$"))
top_weights <- cli::format_inline("{top_weights}")
```

The figure's bottom panels illustrate the effect of the ReLU function, where values less than zero are set to zero. At this point, `r top_weights` have the most substantial effect on the model when judged on the sum of their absolute coefficients. From the contour plot, their patterns are nearly opposite.

However, the model estimates additional parameters that merge these regions, using weights estimated to be most helpful for predicting the data. Those equations were estimated to be: 

```{r}
#| label: nnet-ex-last
#| echo: false
#| results: "asis"

align_equations(
  cbind(coefs$model.2.bias, coefs$model.2.weight),
  c("", paste0("H_{", 1:4, "}")),
  paste0("Y_{", 1:2, "}")
)
```

@fig-nnet-ex-layer-2 illustrates the transition from the hidden layer to the output layer. The weighted sum of the four variables at the top produces the regions of strong positive (or negative) activation for the two classes. Once again, the black curve indicates the zero contour. 

```{r}
#| label: fig-nnet-ex-layer-2
#| dev: "ragg_png"
#| echo: false
#| out-width: 80%
#| fig-width: 7
#| fig-height: 5
#| dev: "ragg_png"
#| warning: false
#| fig-cap: "Heatmaps of the four ReLU units and their output units across the predictor space."
#| fig-alt: "Two rows of panels. The top row shows four heatmaps of ReLU-activated hidden unit values across the predictor space, with zero regions clearly visible. The bottom row shows two output unit heatmaps derived from weighted sums of the hidden units. The two output panels display nearly opposite patterns, with a black contour line at zero marking the class boundary."

output_layer <-
  components |>
  dplyr::select(-starts_with("lp"), -starts_with("relu_")) |>
  pivot_longer(cols = c(starts_with("output"))) |>
  mutate(
    split_up = map(name, ~ strsplit(.x, split = "_")[[1]]),
    name = map_chr(split_up, ~ paste0("Y[", .x[2], "]"))
  ) |> 
  ggplot(aes(A, B)) +
  geom_tile(aes(fill = value)) +
  geom_contour(
    # data = grid_pred,
    aes(z = value),
    breaks = 0,
    col = "black"
  ) +
  facet_wrap(~name, nrow = 1, labeller = label_parsed) +
  scale_fill_gradient2(high = "#420F75FF", low = "#EAC541FF") +
  coord_fixed() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  ) +
  labs(title = "Output", y = NULL)

relu_layer / output_layer
```

The two output panels at the bottom of the figure show different patterns, representing three different regions of the predictor space. The legend in the bottom right clearly shows inappropriate units; we'd like it to reflect probabilities. As previously mentioned, the softmax function coerces these values to be on [0, 1] and sum to one. This produces the class boundary seen in @fig-nnet-ex. 

As previously stated in @sec-fda, ReLU is not a smooth function, so this demonstration will differ somewhat from that for other activation functions; they generally don't completely isolate regions of the predictor space. However, the overall flow of the network remains the same. 

Now that we've discussed the architecture of these models and how the layers relate to one another, let's examine how to estimate model parameters effectively.  

### Training and Optimization {#sec-nnet-training}

Training neural networks can be challenging. Their model structure is composed of multiple nested nonlinear functions that may not be smooth or continuous. The loss function for classification, commonly cross-entropy, is unlikely to be convex as model complexity increases with the addition of more layers. Consequently, we have no guarantees that a gradient-based optimizer will converge. There is also the possibility of local minima that might trap the training process, saddle points that could lead the search in the wrong direction, and regions of vanishing gradients [@hochreiter1998vanishing]. 

First is the challenge of local minima. In some cases, ML models can provide some guarantees regarding the existence of a global minimum loss value^[Generalized linear models are a good example of this]. This may not be the case for neural networks. Their loss surface may contain multiple valleys, where the derivatives of some parameters are zero, but a better solution exists in another region of the parameter space. In other words, the model can appear to converge but has suboptimal performance. @baldi1989neural show that, under some conditions, single-layer feedforward networks can have a unique global minimum. It is expected that as more layers and units are added, the probability of finding a unique optimal solution decreases. When "caught" in a local valley, we need our optimization algorithm to be able to escape. 

A second related issue is saddle points. These are locations where all gradients are zero (i.e., a stationary point), but some directions have increasing loss, while others have decreasing loss. @dauphin2014identifying postulated that these will exist, and their propensity to occur increases with the number of model parameters. The concern here is that, depending on the type of multidimensional curvature encountered, the optimization algorithm may move in the direction of _increasing loss_ or become stuck. 

@fig-stationary has an example with two parameters and a highly nonlinear loss surface. There are four stationary points, the best of which minimizes the loss function at $\boldsymbol{\theta} = [`r round(stationary$x[stationary$type == "minimum"], 1)`, `r round(stationary$y[stationary$type == "minimum"], 1)`]$. The lower-right region has a point associated with the maximum loss. The other two locations are saddle points. The point at $\boldsymbol{\theta} = [`r round(stationary$x[stationary$type == "saddle point" & stationary$x > 6], 1)`, `r round(stationary$y[stationary$type == "saddle point" & stationary$x > 6], 1)`]$ is a good example; moving to the left or right of the point will increase the loss. Moving up or down will reduce the loss (with the higher value being the correct choice within this parameter range). Our optimizer should be able to move in the proper direction or, like simulated annealing, recover and move back to progressively better results. 

::: {#fig-stationary}
::::: {.figure-content}
```{r}
#| label: stationary
#| dev: "ragg_png"
#| echo: false
#| out-width: 60%
#| fig-width: 5.5
#| fig-height: 4.7
#| dev: "ragg_png"
#| fig-alt: "Heatmap of a nonlinear loss surface over two parameters, with contour lines and four labeled stationary points: one minimum, one maximum, and two saddle points. The loss increases from lighter to darker shading. The minimum is in the upper-left region, the maximum in the lower right, and the saddle points lie between them."

loss_grid |>
  rename(loss = res) |>
  ggplot(aes(x, y)) +
  geom_tile(aes(fill = loss)) +
  geom_contour(aes(z = loss), col = "darkred", bins = 7, alpha = 3 / 4) +
  scale_fill_gradient(low = "white", high = "red") +
  scale_shape_manual(values = c(2, 6, 1)) +
  geom_point(
    data = stationary,
    aes(pch = type),
    cex = 2.5,
    col = "black"
  ) +
  coord_fixed(ratio = 5) +
  labs(x = expression(theta[1]), y = expression(theta[2]))
```
:::::

The loss surface for $\psi(\boldsymbol{\theta}) = \theta_1 cos(0.5\,\theta_1) - 2 \exp(-6(\theta_2 - 0.3)^2) + 0.2\, \theta_1 \theta_2$. Four stationary points are shown.
:::

Another possible factor is that many deep learning models are trained on vast amounts of data. This, in itself, imposes constraints on how much data can be held in memory at once, as well as other logistical issues. This may not be an issue for the average application that uses tabular data; however, the problem still persists. 

Before proceeding, recall @sec-gradient-opt where gradient descent was introduced. There are two commonly used classes of gradient-based optimization: 

- _First-order techniques_: $\quad\boldsymbol{\theta}_{i+1} = \boldsymbol{\theta}_i - \alpha\,g(\boldsymbol{\theta}_i)$,

- _Second-order methods_: $\quad\boldsymbol{\theta}_{i+1} = \boldsymbol{\theta}_i - \alpha H(\boldsymbol{\theta_i})^{-1}\,g(\boldsymbol{\theta}_i)$,

where $\boldsymbol{\theta}$ are the unknown parameters, $\alpha$ is the learning rate, $g(\boldsymbol{\theta}_i)$ is the gradient vector of first derivatives, and $H(\boldsymbol{\theta_i})$ is the Hessian matrix of second derivatives. The Hessian measures the curvature of the loss function at the current iteration. Pre-multiplying by its inverse rescales the gradient vector, generally making it more effective. We'll discuss them both in @sec-gradient-descent and @sec-newtons-method. Generally speaking, second-order methods are preferred since they are more accurate _when they are computationally feasible_. However, first-order search is more commonly used because it is faster to compute and can be significantly more memory-efficient.

To demonstrate, we can use a two-layer MLP model for the forestation data. The first layer contains 30 ReLU units and the second has 5 ReLU units, yielding `r num_optimizer_param` model parameters. An L<sub>2</sub> penalty of 0.15 was set to regularize the cross-entropy loss function. A maximum of 100 epochs was specified, but training was instructed to stop after five consecutive epochs with a loss increase (measured using held-out data). An initial learning rate of $\alpha = 0.1$ was used, but a variable, step-wise rate schedule was implemented to modulate the rate across epochs, as described below. 

The model was first trained using basic gradient descent. For example, training was performed 5 times with different initial parameter values. The leftmost panel in @fig-optimizers shows the results. In each case, the loss initially decreased in the first set of epochs. However, in each case, the model used all 100 epochs. This approach was not very successful. There was little reduction in the loss function, and the terminal phase showed only a minuscule decrease, with no meaningful improvement. This _could_ be due to the search approaching the saddle point. In this case, the gradient approaches zero, and first-order gradient searches yield tiny, ineffective updates. We can measure the efficiency of optimizers by computing the time required to complete a full training iteration. For gradient descent, the time was `r round(optimizer_times$per_epoch[optimizer_times$method == "GD"], 3)`s per iteration. 

```{r}
#| label: fig-optimizers
#| echo: false
#| out-width: 70%
#| fig-width: 6
#| fig-height: 2.8
#| warning: false
#| fig-cap: "The loss function, using held-out data, for the same MLP model for four optimization strategies. The lines on each panel represent runs initialized with different random numbers. The profiles are normalized to start from the same initial loss value."
#| fig-alt: "Faceted line plot with four panels comparing optimization strategies: gradient descent, SGD, SGD with momentum, and one additional method. Each panel shows holdout cross-entropy loss over epochs for five runs with different initializations. Gradient descent shows slow convergence with minimal loss reduction. SGD and SGD with momentum converge faster, with some runs stopping early due to early stopping criteria."

optimizers |>
  filter(method != "L-BFGS") |>
  ggplot(aes(epoch, loss_norm, col = seed, group = seed)) +
  geom_line(show.legend = FALSE, linewidth = 3 / 4, alpha = 1 / 2) +
  facet_wrap(~method, scale = "free_x", nrow = 1) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(x = "Epochs", y = "Cross-Entropy (Holdout)")
```

When training neural networks, _stochastic_ gradient descent (SGD) is the standard approach. SGD uses randomly allocated _batches_ of training set samples. The gradient search is updated by processing each batch until the entire training set has been used, which marks the end of that _epoch_. In other words, a potentially large number of parameter updates are computed based on a small number of training point samples. The batch size used here was 64 data points. As a result, with this training set size, about 76 gradient updates are performed before the model has seen all the training data. 

For stochastic gradient descent, we track the process by _epochs_. An epoch is finished when the model has been updated using every data point in the training set. For this analysis, this means there are about 76 parameter updates per epoch. For regular gradient descent and second-order search methods, we'll count their progress as _iterations_ since there are no batches. 

As shown in @fig-optimizers, SGD performs far better than plain GD, and although none of them stop early, there is a relatively flat trend after about 20 epochs, where the search does not substantially reduce the loss. In terms of time per epoch, SGD takes much longer than GD at `r  round(optimizer_times$per_epoch[optimizer_times$method == "SGD"], 3)`s. 

As we'll see in the section below, many extensions to SGD have made it more effective. One improvement is to use _Nesterov momentum_ [@sutskever2013importance]. This initially computes a standard update to the parameters, then approximates the gradient at this position to compute a second prospective step. The actual update that is used is the lookahead position. This technique yields better results than SGD, with several iterations stopping early. However, the training time was longest, using `r round(optimizer_times$per_epoch[optimizer_times$method == "SGD with momentum"], 3)`s per epoch.  

::: {.note-box}
The intended takeaways from this exercise are that training these models can be challenging and that the type of parameter optimization can significantly impact model quality. 
:::

Let's first discuss a few operational aspects of gradient-based methods before turning to first- and second-order search techniques. 

### Initialization {#sec-nnet-initialization}

Network parameter initialization can significantly impact the success of training. Given the size and complexity of these models, it is nearly impossible to analytically derive optimal methodologies for setting initial values. As such, parameters are initialized using random numbers. This means that different initial values will result in different model parameter estimates.

There are a few approaches for simulating initial values.  @lecun2002efficient initialization uses a uniform distribution centered around zero with limits defined by the number of connections feeding into the current node (known as the "fan in" and is denoted as $m_{in}$). They used $U(\pm \sqrt{2/m_{in}})$ as the distribution. There have been a few variations of this: 

- Glorot/Xavier initialization [@glorot2010understanding]: $U(\pm \sqrt{6/(m_{in}+m_{out})})$ 
- He/Kaiming initialization [@he2015delving]: $N(0, \sqrt{2 / m_{in}})$.

where $m_{out}$ is the number of parameters leaving the node. 

One reason that initialization is important is to help mitigate numerical overflow errors. For example, the softmax transformation shown above exponentiates the raw output units (and sums them). If the output units have relatively large values, exponentiating and summing them can exceed allowable numerical tolerances. For example, the output units shown in @fig-nnet-ex-layer-2 have a wide enough range that can be as large as `r max(round(abs(components$output_1)), 0)`. 

### Preprocessing {#sec-nnet-preprocessing}

Neural networks require complete numeric features for computations. 

For non-numeric data, the tools described in @sec-categorical-predictors can be very effective. Additionally, there are pretrained neural network models that can translate the category string to a word embedding [@Boykis_What_are_embeddings_2023], such as  BERT [@devlin2019bert] or GPT-4 [@achiam2023gpt]. The potential upside to doing this is that the resulting numeric features may measure some relative context between categorical values. For example, suppose we have a predictor with levels `red`, `darkred`, `green`, `chartreuse`, and `white`. Basic encoders, such as binary indicators or hashing features, would not recognize that there are three clusters in these values and that sets such as {`red`, `darkred`} and {`green`, `chartreuse`} should have values that are "near" one another. LLMs can measure this semantic similarity and potentially generate more meaningful features. 

If the non-numeric predictors are not complex, a layer could be added early in the network, enabling the model to derive encodings from the data during training. @guo2016entity describes a relatively simple architecture for doing this, which we will see in later sections. While LLMs are better at creating embeddings overall, they are unaware of the _context_ in your data. Estimating the parameters of a simple embedding layer may be more effective, since it is optimized for your dataset. 

For numeric predictors, we suggest avoiding irrelevant predictors and highly collinear feature sets^[Although the need for preprocessing features prior to training a neural network has long been recognized, @lecun2002efficient and @wiesler2014mean re-emphasized that standardization (i.e., centering and scaling) and decorrelating are critical for the optimization algorithm to converge within a reasonable time. ]. @fig-irrelevant-predictors in @sec-feature-selection demonstrated the harm of including predictors unrelated to the outcome in a neural network. Multicollinearity can also affect these models, even if we do not invert the feature matrix or use the Hessian during optimization. Highly redundant predictors can result in the loss function having difficult characteristics, such as ill-conditioned curvature and flat or degenerate regions, as well as a large number of unnecessary parameters. Using an L<sub>2</sub> penalty or a feature extraction method can be very helpful. 

Also, predictors with highly skewed distributions might benefit from a transformation to make their distribution more symmetric. If not, outlying values can exert a significant influence on computations, potentially preventing the generation of high-quality parameter estimates. 

A critical preprocessing technique is to help ensure that only informative predictors are present in the training set. Looking back at @fig-irrelevant-predictors, the simulation results show that including a moderate number of noisy predictors in the training set can substantially reduce performance. In this simulation, performance fell by up to 40% compared with results where all predictors were relevant. 

Finally, since the networks use random initialization, we should ensure that our predictors are standardized to have the same units. If there are only numeric predictors, centering/scaling or orderNorm operations can be very effective^[orderNorm is perhaps more attractive since it standardizes the predictors and coerces them to a symmetric distribution.]. Using this approach, we must also apply the same standardization to binary indicators created from categorical predictors.  lternatively, we could leave 0/1 indicators as-is, but _range_ scale our predictors to also be between zero and one as well. This may have a disadvantage; the range will be common across predictors, but the mean and variance can be quite different. 

### Learning Rates {#sec-nnet-learning-rates}

The gradient descent learning rate ($\alpha$) can profoundly affect model quality, perhaps more than the structural tuning parameters that define complexity (e.g., number of hidden units). A high learning rate ($\approx$ 0.1) will help converge towards effective parameter estimates, but once there, it can cause the model to consistently overshoot the best location. One approach to modulating the rate is to use a scheduler [@lu2022gradient,chapt 4]. These are functions/algorithms that will change the rate (often to decrease it) over the course of epochs. For example, to monotonically decrease the rate, two options are:

- inverse decay: $f(\alpha_0, \delta, i) = \alpha_0/(1 + \delta i)$ 
- exponential decay: $f(\alpha_0, \delta, i) = \alpha_0\exp(-\delta i)$

where $i$ is the epoch number and $\delta$ is a decay value. These two functions gradually decrease the rate so that, in later epochs when we are near an optimum, it is slow enough not to exceed the optimal settings. Another approach is a step function that decreases the rate while maintaining it at a constant level over a range of epochs. Those ranges can be predefined or dynamically adjusted to decrease as the loss function decreases. The function for the former is: 

- step decay: $f(\alpha_0, \zeta, r, i) = \alpha_0 \zeta^{\lfloor i/r\rfloor}$ 

where $\zeta$ is the reduction value, $r$ is the step length, and $\lfloor x\rfloor$ is the floor function. 

Finally, some schedulers modulate the rate up and down. A cyclic schedule [@smith2017cyclical] starts at a maximum value, decreases to a minimum value, and then increases cyclically. For some range $r$ and minimum rate $\alpha_0$, an epoch's location within the series is $cycle = \lfloor 1 + (i / (2r) )\rfloor$ and the rate is

- cyclic: $f(\alpha_0, \alpha, r, i) = \alpha_0 + ( \alpha - \alpha_0 ) \max(0, 1 - x)$

where $x = |(i/r) - (2\,cycle) + 1|$. This is a good option when we are unsure how quickly or slowly convergence will occur. If the rate is too high near an optimum value, it will soon decrease to an appropriate value. Additionally, if the search is stuck in a local optimum (or a saddle point), increasing the learning rate may help it escape. 

@fig-schedulers uses the loss function example from @fig-stationary to demonstrate the effect of rate schedulers. Two starting points are used. One (in orange) is very close to a saddle point, while the other starts from the upper right corner of the parameter space (in green). 

```{r}
#| label: fig-schedulers
#| dev: "ragg_png"
#| echo: false
#| out-width: 85%
#| fig-width: 8
#| fig-height: 2.8
#| dev: "ragg_png"
#| fig-cap: "The effect of different schedulers on the rate and success of convergence."
#| fig-alt: "Faceted heatmap with four panels for constant, cyclic, decay, and stepwise learning rate schedulers. Each panel shows the loss surface in two parameter dimensions with overlaid optimization paths from two starting points. The constant rate path oscillates near the optimum. The cyclic and decay schedules converge more smoothly, while the stepwise schedule shows mixed results depending on starting location."

base_sgd_plot(rates, NULL) + facet_wrap(~schedule, nrow = 1)
```

Using a constant learning rate of $\alpha = 0.1$, the search marches to the optimum value, but oscillates wildly on approach. Ultimately, it falls short of the best estimates because it is constrained to make large jumps at each update. The cyclic schedule allowed rates to move between $\alpha_0 = 0.001$ to $\alpha = 0.1$ with a cycle range of $r = 10$ epochs. This appears to work well, or at least, it had good timing. The decay panel represents inverse decay with $\delta = 0.025$. For the corner starting point, this is a very effective strategy. From the saddle point, we can observe that decreasing the learning rate reduces oscillations as the search progresses and eventually leads to the optimum. Finally, the panel for stepwise reduction used an initial learning rate of $\alpha_0 = 0.1$, $\zeta = 0.5$, and a step length of 20 epochs. This approach worked well when the search began from the saddle location, but performed poorly when starting at the corner. The success depended on when the search was near the optimum; if not timed right, the oscillations were too wide to converge on the best point. A reduction in the step length might help. 

The schedule type and its additional parameter(s) are added to the list of possible values that could be optimized during tuning.

### Regularization {#sec-nnet-regularization}

As seen in @sec-logistic-penalized, we can add a penalty to the loss function to discourage particularly large model coefficients. This is especially beneficial for neural network models. 

While @hanson1988comparing and @weigend1990generalization made early proposals for penalizing the size of estimated parameters, @krogh1991simple proposed using L<sub>2</sub> regularization to improve generalization. This is now a standard approach for training neural networks, and "weight decay" is now synonymous with squared penalties. 

While either L<sub>1</sub>, L<sub>2</sub>, or a mixture of the two can be used, there is some reason to use just L<sub>2</sub> penalization. While either penalty type can mitigate multicollinearity, the quadratic penalty has the advantage of maintaining a smooth loss surface. With an L<sub>1</sub> penalty, we have at least one point on the loss surface that is not differentiable. Also note that using the L<sub>1</sub> penalty will **not** yield exact zeros in the parameter estimates with stochastic gradient descent (as the Lasso model did).

As we will describe later in this chapter, @loshchilov2017decoupled note that there are cases when penalizing the loss function does not precisely match the definition of weight decay (in the neural network context of @hanson1988comparing). Their adjustment for these issues is called the "AdamW" method. 

Another regularization technique is called _dropout_. This method randomly removes some units (hidden or not) from the network. This drops incoming and outgoing connections to other units, which propagates across all layers. These temporary coercions to zero recur with each gradient evaluation. In the final training step, a full set of parameters is used. Dropout has an effect similar to averaging many smaller networks. The dropout rate should be tuned.

### Early Stopping {#sec-early-stopping}

When should training stop? We can estimate the number of epochs that might be needed, but if we make an incorrect guess, we risk under- or overfitting the model. It makes sense to include the number of epochs as a tuning parameter, but this could be costly^[Although the submodel trick described in @sec-submodels could greatly improve the effectiveness of this approach.]. 

If we have some data to spare, an efficient and effective approach is to use early stopping. If we set aside a small proportion of the training set to measure the search's progress, we can set the _maximum_ number of epochs and stop training when the loss worsens (based on our holdout set). Often, we can tune the number of "stopping epochs". If this value were 5, we would stop training after 5 consecutive epochs in which the loss function worsens (and use the parameter estimates from our last best epoch). Values larger than one protect against overreacting when the loss profile is noisy. 

### Batch Learning {#sec-batch-learning}

Operationally, within each epoch of SGD, a random batch is selected to evaluate the gradient and update the parameters accordingly. This occurs for each batch until the entire set is evaluated. This is the end of the epoch. If we are monitoring a holdout set for early stopping, this is the point at which we evaluate the loss function at the current parameter estimate using the holdout set. After this, the process begins again by iterating through the batches. 

SGD is thought to work well because it injects considerable variation into the gradients and updates. This would generally be considered a bad thing, but, as with simulated annealing, it makes the optimization less greedy. The somewhat noisy directions the optimizer takes can help the optimization escape from local optimums and be more resilient when the search veers near a saddle point or other regions where the gradient may be flat or consistently zero. @hoffer2017train referred to SGD as a "high-dimensional random walk on a random potential" process. 

As the search proceeds, lowering the learning rate can mitigate the extra noise in the gradients so that, hopefully, the search does not move away from the best parameter estimates. 

We can tune the batch size if there is no intuition for what a good value might be. Smallish batch sizes could range from 16 to 512 training set points^[Typically on log<sub>2</sub> units], depending on the training set size. @keskar2016large remarks that smaller batches tend to explore the parameter space more thoroughly and that, as the batch size becomes larger, SGD will find a solution that is closer to the initial values. 

@loshchilov2017decoupled make a few interesting observations related to L<sub>2</sub> penalization with stochastic gradient descent. Their results show that as the number of batches increases, the optimal penalty decreases. In other words, as the batch size decreases, the amount of penalization should also decrease. 

In practice, the batch size is greatly influenced by the need to balance execution time and hardware memory limitations. 

### Gradient Descent Algorithms {#sec-gradient-descent}

Gradient descent methods for training neural networks have become a popular research topic over the last 15 years. We'll discuss a few major techniques here, but there are myriad other tools for training these models. In this section, we'll temporarily forgo discussing changes from epoch to epoch. Since we are focused on SGD, we'll use the term "update" to refer to the point at which the parameters move to a different location after a batch step. 

#### Momentum {.unnumbered}

To start, let's consider a form of momentum in gradient search^[In @sec-nnet-training, we previously described Nesterov momentum. This is unrelated to the current tool bearing the same name.]. Momentum is a useful tool that can help in situations such as the results in @fig-schedulers, where updates cause the search path to take large, ineffective jumps. When the learning rate is consistently large, the search path bounces back and forth because the gradients are large in magnitude but point in different directions at each step. 

Momentum is one of several tools to adjust the gradient vector based on previous gradients. For iteration $i$, a vector is computed that is a combination of the current and previous gradients^[For the first update, the value is simply $v_1 = \alpha g(\boldsymbol{\theta}_i)$.]: 

$$
v_i = \phi v_{i-1} + \alpha_i g(\boldsymbol{\theta}_i)
$$

and this is used to make the update $\boldsymbol{\theta}_{i+1} = \boldsymbol{\theta}_i - v_i$. This weighting system emphasizes recent gradients while also accounting for previous gradients. This can accelerate the search and diminish oscillations.   

Values of $\phi$ typically range from 0.8 to 0.99, representing a weighted average of previous gradient directions, with greater emphasis on recent iterations as $\phi$ increases. If the directions have been fairly stable, there is little change; however, if they are oscillating, the direction takes a more consistent central route with fewer directional jumps. It also helps if the search is in a region where gradients are close to zero; historical gradients will have some non-zero values that are unlikely to become stuck with updates near zero. 

@fig-sgd shows the results when momentum is used for our 2D toy example^[Since the loss function for this example is based on a single, deterministic value, the momentum value was set very low ($\phi = 2/3$). However, the overall trend is consistent with results from a typical performance metric like cross-entropy.]. The cyclic rate settings were used in this example, and the results are very similar to those from the previous path when initialized near the saddle point. The corner-point starting value shows that momentum can lead to a location near the optimal value, but takes a more circuitous route. 

```{r}
#| label: fig-sgd
#| dev: "ragg_png"
#| echo: false
#| out-width: 85%
#| fig-width: 8
#| fig-height: 2.8
#| dev: "ragg_png"
#| warning: false
#| fig-cap: "Examples of four different optimizers for the loss surface previously shown in @fig-schedulers."
#| fig-alt: "Faceted heatmap with four panels for Momentum, AdaGrad, RMSprop with Nesterov, and ADAM optimizers. Each panel shows optimization paths from two starting points overlaid on the same two-dimensional loss surface. Momentum and ADAM converge effectively toward the minimum. AdaGrad stalls before reaching the optimum due to the accumulation of squared gradients. RMSprop with Nesterov converges smoothly."

bind_rows(momentum_res, adam_res, rms_prop_res, adagrad_res) |>
  mutate(
    technique = factor(
      technique,
      levels = c("Momentum", "AdaGrad", "RMSprop + Nesterov", "ADAM")
    )
  ) |>
  base_sgd_plot(NULL) +
  facet_wrap(~technique, nrow = 1)
```

Momentum should be used in conjunction with a rate scheduler, or at the very least, without a consistently large learning rate. Making very large jumps in a central direction might lead the search to points far away from regions of improvement. Penalization also helps. 

#### Adaptive Learning Rates {.unnumbered}

There is also a class of gradient descent methods that try to avoid using a single learning rate for all of the elements of $\boldsymbol{\theta}$.  Instead, they use a constant learning rate and, at each update, adjust the rate using functions of the gradient vector so that the rates are different for each $\theta_j$. 

One approach is the adaptive gradient algorithm, a.k.a. AdaGrad [@duchi2011adaptive], that scales the constant learning rate $\alpha$ by the accumulated square of the gradients: 

$$
\boldsymbol{v}_i =\sum_{b=1}^t g(\boldsymbol{\theta}_b)^2 
$$
so that 

$$
\boldsymbol{\theta}_{i+1} = \boldsymbol{\theta}_i - \frac{\alpha}{\sqrt{\boldsymbol{v}_i + \epsilon}}\, g(\boldsymbol{\theta}_i)
$$

where $\epsilon$ is a small constant to avoid division by zero. The square root in the denominator is required to keep the denominator units the same as those in the numerator. AdaGrad has the advantage of keeping the adjusted learning rate steady for elements of $g(\boldsymbol{\theta})$ that do not change substantially from update to update. At update $i$, the contribution to the learning rate is inversely proportional to the squared gradient value; we want to go far when the slope is large.

The problem is the _unweighted_ accumulation of squared values. For a reasonable number of updates, elements of $\boldsymbol{v}$ can explode to very large values, and this prevents almost _any_ movement of $\boldsymbol{\theta}$ from update to update. For the example shown in @fig-sgd, some of these values were in the thousands, and progress clearly halted before reaching the optimal parameters. 

One method of fixing this problem is RMSprop [@hinton2012neural]. Here, the denominator uses an exponentially weighted moving _average_ of the squared gradients: 

$$
\boldsymbol{v}_i = \rho\, \boldsymbol{v}_{i-1} + (1 - \rho)\, g(\boldsymbol{\theta}_i)^2 
$$

This formulation places more weight on the current and recent gradients than the statistic used by AdaGrad. The RMSprop update becomes: 

$$
\boldsymbol{\theta}_{i+1} = \boldsymbol{\theta}_i - \frac{\alpha}{\sqrt{\boldsymbol{v}_i + \epsilon}}\, g(\boldsymbol{\theta}_i).
$$

The "RMS" in the name is due to taking the square root of the weighted sum of squared gradients (i.e., similar to a traditional root mean square). 

@goodfellow2016deep suggested using RMSprop with Nesterov momentum. This leads to a more complicated procedure. First, we compute our initial update: $\quad\tilde{\boldsymbol{\theta}}_{i} = \boldsymbol{\theta}_{i} - \phi\,\boldsymbol{\delta}_{i-1}$ where $\boldsymbol{\delta}_0$ is initialized with a vector of zeros. We compute the gradient at this temporary update and use this to compute $\boldsymbol{v}_i$. From here: 

$$
\boldsymbol{\delta}_{i} = \phi\,\boldsymbol{\delta}_{i-1} - \frac{\alpha}{\sqrt{\boldsymbol{v}_i + \epsilon}}\, g(\tilde{\boldsymbol{\theta}}_i)
$$

The results are in @fig-sgd. The curves are similar to those for AdaGrad except that they do reach the location of minimum loss. The path for the starting value near the saddle point exhibits several minor oscillations en route to the optimal result. 

The adaptive moment estimation (Adam) optimizer from @kingma2017adammethodstochasticoptimization  extends the moving average approach to the gradient term as well:  

$$
\begin{align}
\boldsymbol{m}_i &= \phi\, \boldsymbol{m}_{i-1} + (1 - \phi)\, g(\boldsymbol{\theta}_i) \notag \\
\boldsymbol{v}_i &= \rho\, \boldsymbol{v}_{i-1} + (1 - \rho)\, g(\boldsymbol{\theta}_i)^2 \notag 
\end{align}
$$

Adam also tries to normalize these quantities so that the first few updates have larger values:   

$$
\begin{align}
\boldsymbol{m}^*_i &= \frac{\boldsymbol{m}_i}{1 - \phi^i}\notag \\
\boldsymbol{v}^*_i &= \frac{\boldsymbol{v}_i}{1 - \rho^i} \notag 
\end{align}
$$

This helps offset the initialization of $\boldsymbol{m}_0 = \boldsymbol{v}_0 = \boldsymbol{0}$. At update $i$, Adam's proposal is

$$
\boldsymbol{\theta}_{i+1} = \boldsymbol{\theta}_i - \frac{\alpha}{\sqrt{\boldsymbol{v}^*_i + \epsilon}}\,\boldsymbol{m}^*_i.
$$

The results above show that Adam's path towards the optimal parameters is slightly wobbly compared to RMSprop but otherwise effective. 

Both $\phi$ and $\rho$ are tunable in these models and typically range within [0.8, 0.99]. The constant learning rate $\alpha$ is also tunable. In @fig-sgd, $\phi=\rho=0.9$. AdaGrad, Adam, and RMSprop used $\alpha = 0.1$. For RMSprop, decreasing the learning rate to $\alpha = 0.08$ removed the oscillations shown in @fig-sgd.

As alluded to earlier, @loshchilov2017decoupled noted that, for Adam, penalizing the loss function does not align with the definition of weight decay. Without a remedy, they found that Adam is not as effective as it could be. Their method, called AdamW, includes the penalty in the gradient definition. The gradient used to compute $\boldsymbol{m}_i$ and $\boldsymbol{v}_i$ is a combination of the previous "raw" gradient and the penalty ($\lambda$): 

$$
g(\boldsymbol{\theta}_i) = \nabla \psi(\boldsymbol{\theta}_{i}) - \lambda \boldsymbol{\theta}_{i}
$$

The updating equation also uses an extra term containing the penalty: 

$$
\boldsymbol{\theta}_{i+1} = \boldsymbol{\theta}_i - \alpha_i\left[\frac{\boldsymbol{m}^*_i}{\sqrt{\boldsymbol{v}^*_i + \epsilon}} + \lambda \boldsymbol{\theta}_{i}\right].
$$

Note that the learning rate $\alpha$ now has a subscript. They also recommend trying a cyclic learning rate scheduler with AdamW.  

AdamW has become very popular for training and has superseded Adam. 

#### Other Methods {.unnumbered} 

As previously mentioned, there are numerous other variations of SGD that will not be discussed here, including AdaMax [@Diederik2015], Nadam [@dozat2016], MadGrad [@defazio2021], and others.  

### Newton's Method {#sec-newtons-method}

As previously shown, Newton's method (a.k.a., the Newton--Raphson method) uses both gradients and the Hessian matrix to update parameters. This approach can be very effective when the local loss surface can be approximated by a second-degree polynomial. When the local search includes a stationary point, the signs of the eigenvalues of the Hessian can characterize it: all non-negative imply a maximum, all non-positive imply a minimum, and mixed signs indicate a saddle point. 

One challenge for Newton's Method concerns the Hessian. If there are $p$ parameters, the matrix needs to save $p(p+1)/2$ parameters. For neural networks, $p$ can be in the millions, so the basic application of Newton's method can be slow and might require infeasible amounts of memory (depending on the model size). We also need to invert the Hessian, making it even more costly.

We can approximate Newton's method using the Limited-memory Broyden--Fletcher--Goldfarb--Shanno (L-BFGS) search algorithm [@lu2022gradient,chapt 6]. It uses past gradients to approximate the Hessian as the search proceeds. L-BFGS computes gradients at different values of $\boldsymbol{\theta}$ for more than just updating the parameters (i.e., Hessian approximation and the line search), so it is not a good solution when loss function or gradient evaluations are expensive. 

However, if we have a small dataset, a low-to-medium-complexity architecture, or both, L-BFGS can be a very effective tool for parameter estimation. 

### Forestation Modeling {#sec-nnet-forestation}

```{r}
#| label: load-forested-mlp
#| include: false

# Code in R/forested_mlp.R
load("../RData/forested_mlp.RData")

brier_param_cor <-
  cor(brier_and_params$mean, brier_and_params$num_param, method = "spearman") |>
  round(2)

num_metrics <- vctrs::vec_unique_count(mlp_ranks$.metric)
num_configs <- nrow(mlp_ranks) / num_metrics
num_grid <- vctrs::vec_unique_count(mlp_ranks$.config)

wide_mtr <-
  mlp_ranks |>
  dplyr::select(wflow_id, .config, .metric, mean) |>
  pivot_wider(
    id_cols = c(wflow_id, .config),
    names_from = .metric,
    values_from = mean
  )
mtr_cor <- cor(wide_mtr$brier_class, wide_mtr$roc_auc) |> round(2)
```

For the forested data, there are a few options for preprocessing: 

- Use binary indicators to represent the `r length(levels(forested_train$county))` counties or a supervised effect encoding strategy. 
- Several predictors are highly skewed. We can leave them as-is before centering/scaling, or use an orderNorm transformation to coerce each predictor to a standard normal distribution. 
- Deal with the high correlation for some predictors via PCA extraction or leave as-is in the data. 

Several combinations of these operations were used: 

- An effect encoding with and without PCA feature extraction.
- Standard binary indicators with and without PCA feature extraction and with and without a symmetry transformation (i.e., orderNorm versus centering/scaling). 

For the PCA extraction, all possible components were used, and the entire predictor set was renormalized after PCA. 

For the model, there are a fair number of tuning parameters to combine: 

- One or two layers.
- The number of hidden units. For each layer, these ranged from two to fifty units. 
- The activation type: ReLU, elu, tanh, and tanhshrink.
- L<sub>2</sub> regularization values between 10<sup>-10</sup> and 10<sup>-1</sup>.
- Batch sizes between 2<sup>3</sup> and 2<sup>8</sup>.
- Learning rates between 10<sup>-4</sup> and 10<sup>-1</sup>.
- Learning rate schedules: constant, cyclic, and time decay.
- Momentum values between [0.8, 0.99].

AdamW was used to train the models, with a maximum of 25 epochs, but early stopping could occur after 5 consecutive epochs of poor performance. 

A space-filling design with `r num_grid` candidates was evaluated. The six preprocessing schemes listed above were each run with the same model grid, resulting in `r num_configs` candidates to be evaluated. 

@fig-forest-mlp-ranks shows the results for the `r num_configs` candidates. They are ranked by their resampling estimate of the Brier scores, and the figure presents these rankings from best to worst, breaking out the combinations of preprocessing scenarios. While each panel contains models with sufficiently low Brier scores, several trends are evident. The "Indicators" panel represents models that use dummy-variable indicators with no symmetry transformation, and they did not perform well (with or without PCA feature extraction). 

```{r}
#| label: fig-forest-mlp-ranks
#| echo: false
#| out-width: 95%
#| fig-width: 9
#| fig-height: 5.2
#| fig-cap: Rankings of multilayer perceptron models for the forestation data using the Brier score. The panels represent different types of preprocessing, and the vertical lines are 90% confidence intervals.
#| fig-alt: "Faceted dot plot of MLP model rankings by Brier score, arranged in a grid with columns for encoding type (Indicators, Indicators plus Symmetry, Effect Encoding plus Symmetry) and rows for PCA versus no PCA. Points are distinguished by one-layer and two-layer architectures. Vertical lines show 90% confidence intervals. PCA preprocessing generally produces worse results. The best performance comes from Indicators plus Symmetry and Effect Encoding plus Symmetry without PCA."

mlp_ranks |>
  mutate(
    encoding = map_chr(wflow_id, ~ strsplit(.x, split = "_")[[1]][1]),
    encoding = case_when(
      encoding == "orderNorm" ~ "Indicators + Symmetry",
      encoding == "plain" ~ "Indicators",
      .default = "Effect Encoding + Symmetry"
    ),
    encoding = factor(
      encoding,
      levels = c(
        "Indicators",
        "Indicators + Symmetry",
        "Effect Encoding + Symmetry"
      )
    ),
    extraction = map_chr(wflow_id, ~ strsplit(.x, split = "_")[[1]][2]),
    extraction = ifelse(extraction == "pca", "PCA", "No PCA"),
    extraction = factor(extraction, levels = c("No PCA", "PCA")),
    layers = ifelse(grepl("2L", wflow_id), "2 layers", "1 layer"),
    layers = factor(layers, levels = c("1 layer", "2 layers")),
    lower = mean - qnorm(0.05) * std_err,
    upper = mean + qnorm(0.05) * std_err
  ) |>
  filter(.metric == "brier_class") |>
  ggplot(aes(x = rank, col = layers)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), alpha = 1 / 3) +
  geom_point(aes(y = mean), cex = 1 / 2) +
  facet_grid(extraction ~ encoding) +
  scale_color_manual(values = c("#0076BBFF", "#E9A17CFF")) +
  labs(x = "Overall Rank", y = "Brier Score") +
  theme(legend.position = "top")
```

In fact, PCA has suboptimal results in each of the three bottom panels. This could be due to the components being unsupervised; they are not optimized for prediction. Alternatively, the last few components might be measuring noise and have less relevance to the model. Adding irrelevant predictors to neural networks can diminish their effectiveness. 

The two cases that did the best were the "Indicators + Symmetry" and "Effect Encoding + Symmetry" panels. While both used the orderNorm transformation, the former panel employed indicators, whereas the latter used supervised effect encoding, with a single column representing the county effect. This suggests that eliminating distributional skew in the predictors may yield a modest improvement in predictive performance. 

```{r}
#| label: tbl-mlp-top-five
#| tbl-cap: "The best models as ranked by the Brier score."

top_five <-
  mlp_ranks |>
  filter(.metric == "brier_class") |>
  slice_min(mean, n = 5)

top_five |>
  inner_join(
    brier_and_params |> distinct(wflow_id, .config, num_param),
    by = join_by(wflow_id, .config)
  ) |>
  dplyr::select(wflow_id, .config, rank, num_param) |>
  inner_join(mlp_best_mtr, by = join_by(wflow_id, .config)) |>
  mutate(
    tmp1 = map2_chr(hidden_units, activation, ~ paste0(.x, " ", .y)),
    tmp2 = map2_chr(hidden_units_2, activation_2, ~ paste0(", ", .x, " ", .y)),
    tmp2 = ifelse(is.na(hidden_units_2), "", tmp2),
    Architecture = map2_chr(tmp1, tmp2, ~ paste0(.x, " ", .y)),
    Penalty = format(log10(penalty), digits = 3),
    Penalty = paste0("10<sup>", Penalty, "</sup>"),
    rate_schedule = ifelse(
      rate_schedule == "decay_time",
      "inverse decay",
      rate_schedule
    ),
    learn_rate = format(log10(learn_rate), digits = 3),
    learn_rate = paste0("10<sup>", learn_rate, "</sup>"),
    learn_rate = map2_chr(
      learn_rate,
      rate_schedule,
      ~ paste0(.x, " (", .y, ")")
    ),
    encoding = map_chr(wflow_id, ~ strsplit(.x, split = "_")[[1]][1]),
    encoding = case_when(
      encoding == "orderNorm" ~ "Indicators, Symmetry",
      encoding == "plain" ~ "Indicators",
      .default = "Effect Encoding, Symmetry"
    ),
    extraction = map_chr(wflow_id, ~ strsplit(.x, split = "_")[[1]][2]),
    extraction = ifelse(extraction == "pca", "PCA", "No PCA"),
    extraction = factor(extraction, levels = c("No PCA", "PCA")),
    Preprocessing = map2_chr(encoding, extraction, ~ paste0(.x, ", ", .y)),
    across(where(is.numeric), ~ format(.x, digits = 3))
  ) |>
  # fmt: skip
  dplyr::select(-wflow_id, -.config, -hidden_units, -hidden_units_2, -tmp1, -tmp2,
         -activation, -activation_2, -.estimator, -n, -std_err, -rate_schedule, 
         -encoding, -extraction, -penalty, -rank) |>
  # fmt: skip
  rename(`Learning Rate` = learn_rate, 
         `Batch Size` = batch_size, 
         Momentum = momentum, 
         `# Parameters` = num_param,
  ) |>
  pivot_wider(
    id_cols = c(
      `Learning Rate`,
      `Batch Size`,
      Momentum,
      Architecture,
      Penalty,
      Preprocessing,
      `# Parameters`
    ),
    names_from = c(.metric),
    values_from = c(mean)
  ) |>
  dplyr::select(
    Preprocessing,
    Architecture,
    `# Parameters`,
    Penalty,
    `Batch Size`,
    `Learning Rate`,
    Momentum,
    Brier = brier_class,
    ROC = roc_auc
  ) |>
  gt() |>
  fmt_markdown(columns = c(Penalty, `Learning Rate`)) |>
  tab_spanner("Performance", columns = c(Brier, ROC)) |>
  fmt_number(decimals = 3, columns = c(Brier, ROC)) |>
  cols_width(
    Preprocessing ~ pct(20),
    Architecture ~ pct(11),
    `# Parameters` ~ pct(9),
    Penalty ~ pct(9),
    `Batch Size` ~ pct(9),
    `Learning Rate` ~ pct(15),
    Momentum ~ pct(10),
    Brier ~ pct(8),
    ROC ~ pct(8)
  )
```

Digging in further, @tbl-mlp-top-five displays the top five candidates from a pool of 300. They are a mixture of one- and two-layer networks, but note that we have to go to the third decimal place to distinguish their Brier scores. Constant learning rates performed well as long as the rates were fairly low and the batch sizes were fairly small. The number of parameters also varied; increasing the model's complexity did not appear to help on this dataset. In fact, the rank correlation between the Brier score and the number of parameters was `r brier_param_cor`, indicating little evidence that adding more layers and/or units to the MLP improves the model on these data.

The Brier scores are low, and the areas under the ROC curves are all at least 0.94, indicating that the predictions effectively separate the classes and that the probability estimates are accurate. Across the candidates, the rank correlation between the Brier score and the ROC AUC was `r mtr_cor`; in this case, their optimization results are well-aligned. 

```{r}
#| label: mlp-mtr-diffs
#| include: false
#| eval: false

mtr_differences <- function(split, ref_model) {
  mtr_vals <-
    split |>
    rsample::analysis() |>
    pivot_longer(cols = c(-class), names_to = "Model", values_to = "value") |>
    group_by(Model) |>
    cls_mtr(class, value) |>
    dplyr::select(-.estimator) |>
    pivot_wider(
      id_cols = c(.metric),
      names_from = Model,
      values_from = .estimate
    ) |>
    full_join(metrics_info(cls_mtr), by = ".metric") |>
    dplyr::select(-type)
  models <- colnames(mtr_vals)
  models <- setdiff(models, c(".metric", "direction", ref_model))
  for (i in models) {
    diffs <- mtr_vals[[ref_model]] - mtr_vals[[i]]
    new_nm <- paste0("loss_", i)
    mtr_vals[[new_nm]] <-
      if_else(mtr_vals$direction == "minimize", -diffs, diffs)
  }
  if (length(models) > 1) {
    mtr_vals <-
      mtr_vals |>
      dplyr::select(.metric, starts_with("loss_")) |>
      pivot_longer(
        cols = c(starts_with("loss_")),
        names_to = "term",
        values_to = "estimate"
      ) |>
      mutate(term = gsub("loss_", "", term))
  } else {
    mtr_vals <-
      mtr_vals |>
      dplyr::select(.metric, all_of(new_nm)) |>
      set_names(c(".metric", "estimate")) |>
      mutate(term = i)
  }
  mtr_vals
}

# ------------------------------------------------------------------------------

nonlin_pred <-
  mlp_best_res |>
  collect_predictions(parameters = mlp_best_config) |>
  dplyr::select(.row, mlp = .pred_Yes) |>
  full_join(
    forest_knn_pred |> dplyr::select(.row, class, knn = .pred_Yes),
    by = ".row"
  ) |>
  dplyr::select(-.row)

set.seed(789)
mlp_mtr_bt <-
  nonlin_pred |>
  bootstraps(times = 2000) |>
  mutate(stats = map(splits, mtr_differences, ref_model = "mlp"))

mlp_mtr_diffs <-
  mlp_mtr_bt |>
  int_pctl(stats, alpha = 0.1)

mlp_mtr <-
  mlp_best_res |>
  collect_metrics() |>
  inner_join(mlp_best_config |> dplyr::select(.config), by = ".config")

mlp_brier <- mlp_mtr$mean[mlp_mtr$.metric == "brier_class"]
knn_brier <-
  forest_knn_res |>
  collect_metrics() |>
  filter(.metric == "brier_class") |>
  inner_join(
    forest_knn_best,
    by = join_by(neighbors, weight_func, dist_power, prop_terms, .config)
  ) |>
  pluck("mean")

mlp_roc <- mlp_mtr$mean[mlp_mtr$.metric == "roc_auc"]
knn_roc <-
  forest_knn_res |>
  collect_metrics() |>
  filter(.metric == "roc_auc") |>
  inner_join(
    forest_knn_best,
    by = join_by(neighbors, weight_func, dist_power, prop_terms, .config)
  ) |>
  pluck("mean")
```

`r r_comp("cls-neural-nets.html#sec-cls-nnet")`

## Modern Tabular Network Models {#sec-cls-tab-net}

The previous section describes classical neural networks for tabular data. However, there has been increasing research focus on adapting modern network components designed for image or text data to improve tabular data models. 

A fair amount of research has been motivated by the observation that tree ensembles often outperform tabular neural network models and are generally lower maintenance. As such, some methodologies try to add direct analogs to tree operations. In other cases, tabular models incorporate the same ideas that have catapulted neural networks to greater effectiveness in generative AI: attention and in-context learning. This section will describe a handful of potential upgrades to neural networks to improve their performance on tabular data.

@gorishniy2021revisiting, @borisov2022deep, and @shwartz2022tabular have good summaries of available models. We'll describe several here that highlight some of the current trends. 

### In-Network Normalization {#sec-in-network-normalization}

We previously noted that substantial preprocessing is required for our original predictors. @wiesler2014mean notes that the effects of this initial preprocessing can vanish as more layers are added to the network; there is no guarantee that the hidden units in later layers will have distributional properties that encourage convergence. 

Consider the example in  @sec-nnet-activation where, for a basic MLP model with ReLU activation, large portions of the parameter space have values of absolute zero. No matter the distribution entering the layer, the outputs are likely to have a large point mass at zero and perhaps a second distributional mode. 

For example, @fig-nnet-hidden-dist shows the distribution of a specific hidden unit in a model from @fig-nnet-ex, albeit with a larger batch size, showing how it changes over batches and epochs. The red point shows how often absolute zero values occur. Within a batch, the point mass at zero can fluctuate over epochs. Given that the activated values must be non-negative, the distributions can become fairly skewed over epochs, although two of the distributions are somewhat bell-shaped for non-zero values. In any case, the values have certainly deviated from the centered and scaled distributions of the original predictor values.  

```{r}
#| label: fig-nnet-hidden-dist
#| echo: false
#| out-width: 90%
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| fig-cap: "Histograms of the distributions of a hidden unit with ReLU activation across batches and epochs. The red point indicates the frequency with which absolute zero values occur. The model and data are from @fig-nnet-ex but with a larger batch size (225 instead of 16)."

load("../RData/hidden_units_trace.RData")

vert_format <-
  hidden_units_data |>
  pivot_longer(
    cols = c(starts_with("hidden_")),
    names_to = "embedding",
    values_to = "values"
  ) |>
  filter(epoch %in% c(1, 25, 50, 75, 100) & embedding == "hidden_2")

zeros <- vert_format |>
  filter(values > -.Machine$double.eps & values < .Machine$double.eps)|> 
  count(epoch, batch, embedding) |> 
  mutate(values = 0)
non_zero <- vert_format |>
  filter(values > .Machine$double.eps | values < -.Machine$double.eps)

non_zero |>
  ggplot(aes(values)) +
  geom_histogram(col = "white", binwidth = 0.06) +
  geom_point(data = zeros, aes(y = n), col = "red", alpha = 1 / 2, cex = 1) +
  geom_segment(data = zeros, aes(x = 0, xend = 0, y = 0, yend = n), 
               col = "red", alpha = 1 / 2) +
  facet_grid(epoch ~ batch, labeller = "label_both") +
  labs(x = "Hidden Unit Value", y = "Count")
```

@wiesler2014mean suggested that standardizing the embeddings at different points within the network could yield the same benefits for the model as initial standardization. Their solution was to use batches during SGD to compute a running mean of the model's internal features and to center the embeddings. 

@ioffe2015batch took this idea a few steps further to minimize _covariate shifts_ between batches. Their approach, called _batch normalization_ (a.k.a. "BatchNorm"), standardizes the current embeddings with some extra parameters:

$$
\tilde{H}_J = b\left(H_j\right) = \frac{H_j - \bar{H_j}}{\sqrt{s_j^2 + \epsilon}}\,  \theta_1 + \theta_2
$$

where $\epsilon$ is a small constant and $\bar{H_j}$ and $s_j^2$ are the estimated mean and variance of embedding $j$, respectively. Recall that during SGD, small batches of the training set are used. The sample mean and variance are computed on the current mini-batch of data, and a moving average of their values is maintained. The extra term $\theta_1$ estimates a scale shift and $\theta_2$ estimates a location shift. These are learned parameters; they are estimated via gradient descent. After training, new samples being predicted are normalized using moving-average statistics based on the means and variances produced in the min-batches. 

In @ioffe2015batch, batch normalization occurs between the formation of a new linear predictor transformation and the application of a nonlinear activation. For example, this would have occurred between the two rows of variables shown in @fig-nnet-ex-layer-1. However, @he2016deep showed that "pre-activation" works better for their model. This means normalization occurs before forming the linear combination, which is also what @gorishniy2021revisiting suggests for tabular data. 

BatchNorm doesn't increase the model's predictive capacity; it generally lowers the difficulty level of SGD. It is unclear how. There is some dispute as to whether the correction for covariate shifts is the reason for its effectiveness. There is some belief that BatchNorm smooths the objective function [@santurkar2018does;@peng2023re]. 

Two other in-network normalizations to mention are LayerNorm [@ba2016layer] and RMSNorm [@zhang2019root]. Both of these normalize across the embeddings per sample. For example, LayerNorm computes a mean and standard deviation across features for a single row. RMSNorm also operates across features, but divides them by the square root of the sum of the squared values. Note that if the features have a distribution centered around zero, this is the same as the spatial-sign transformation [@eq-spatial-sign] that projects the data to a hypersphere of radius $\sqrt{p}$ (instead of 1.0). As such, RMSNorm imparts some robustness against unusually large feature values and constrains the range of the normalized values. 

### Residual Networks {#sec-cls-resnet}

Residual networks, a.k.a. ResNet, modify the basic MLP architecture by adding skip connections: the input to each block is added directly to the block's output before passing to the next layer^[Except for the final layer.]. 

Let's illustrate this using an MLP where the number of hidden units in each of the three layers is the same as the number of predictors ($p$), as shown on the left:  

$$
\overbrace{
\begin{aligned}
\boldsymbol{H}_1 &= f_1(\boldsymbol{x}) \notag \\
\boldsymbol{H}_2 &= f_2(\boldsymbol{H}_1) \notag \\
\boldsymbol{H}_3 &= f_3(\boldsymbol{H}_2) \notag \\
\boldsymbol{Y} &= f_4(\boldsymbol{H}_3) \notag
\end{aligned}
}^{\text{Standard MLP}}
\qquad\qquad
\overbrace{
\begin{alignedat}{3}
\boldsymbol{H}_1 &{}={} & f_1(\boldsymbol{x}) &{}+{} &\boldsymbol{x} \notag \\
\boldsymbol{H}_2 &{}={} & f_2(\boldsymbol{H}_1) &{}+{} &\boldsymbol{H}_1 \notag \\
\boldsymbol{H}_3 &{}={} & f_3(\boldsymbol{H}_2) &{}+{} &\boldsymbol{H}_2 \notag \\
\boldsymbol{Y} &{}={} & f_4(\boldsymbol{H}_3) \notag
\end{alignedat}}^{\text{Residual Connections}}
$$  {#eq-resid-net}

For the MLP architecture, the idea is that ordinary feed-forward networks are tasked with learning, at each layer, a better representation of the outcome via $f_i(\cdot)$. This parameterization may be difficult to optimize because changes between layers may be very small, leading to potential issues with gradient magnitudes.  

A residual network, shown on the right-hand side of @eq-resid-net, takes the current features prior to the hidden layer and adds them back into the embeddings (post-activation). In other words, the skip comes *before* $f_i(\cdot)$ and is added to the result *after* $f_i(\cdot)$ applies its activation. We'll use the term _residual block_ to describe the operations shown in each of the top three rows of the right-hand side.

 By adding the previous embeddings to this function, we've reframed the problem so that we are really learning $f_i(\boldsymbol{H}_{i-1}) = \boldsymbol{H}_i - \boldsymbol{H}_{i-1}$; a residual. 

This process makes training easier, especially for models with many layers. Early practitioners assumed that performance could be increased simply by adding more layers. This was not the case. @he2016deep showed that large models paradoxically saw both training and test error increase as the number of layers increased. This isn't believed to be related to overfitting, but due to how gradients flow through the system during optimization. Residual connections address this by giving gradients a direct path back through the network: for layer $i$, the gradient is $f_i'(\boldsymbol{H}_{i-1}) + I$. If we add too much capacity to the model (e.g., more layers), it can simply estimate zero for the residual. Also, the extra term $I$ in the gradient prevents them from becoming too small across the network, even if $f_i'(\boldsymbol{H}_{i-1})$ has elements that are approximately zero. We'll describe other potential benefits to this model towards the end of this section. 

@he2016deep is generally credited for the ResNet model. It showed significant improvements for convolutional neural networks for image analysis.  For tabular data, we'll focus on the formulation in Section 3.2 of @gorishniy2021revisiting. With this model, we'll refer to the elements being added as the "skip path" and the other components (e.g., nonlinear activation) as the "main path". 

Let's get back to the particulars. Our equations on the left-hand side of @eq-resid-net are unrealistic, as they assume the embedding dimension remains the same size as the predictor dimension throughout the network. Once we loosen that constraint, additional considerations are required to align the dimensions

Suppose we begin with $p=2$ predictors within $\boldsymbol{x}$. Suppose we use a "bottleneck" architecture in which the hidden layers have output dimensions (10, 3, 10). If $\boldsymbol{H}_1$ is (10 × 1) and $\boldsymbol{x}$ is (2 × 1), how can we add those to one another?

The answer is that we use a parameter matrix to form a linear combination of the inputs, thereby achieving the correct dimensionality (i.e., projecting the inputs to the correct dimensions). For the first hidden layer, the skip path adds $\boldsymbol{P}_1\boldsymbol{x}$ where $\boldsymbol{P}_1$ is a (10 × 2) _projection matrix_. This generates a version of the original predictors that has the correct size. 

We also need to project the inputs again for the main path, where activation is applied. We'll use a different matrix of parameters ($\boldsymbol{W}_1$) that is the same size as $\boldsymbol{P}_1$. Both matrices coerce the input to the correct dimension, but serve different purposes:

- $\boldsymbol{P}_1\boldsymbol{x}$ should represent the original input as plainly as possible in order to produce an appropriate residual effect. 
- $\boldsymbol{W}_1^{(1)}\boldsymbol{x}$ is designed and estimated to support the nonlinear activation via $f_1(\bullet)$ so that it can learn better outgoing embedding. 

The elements of both of these matrices are learnable. The first residual block is: 

$$
\boldsymbol{H}_1 = \boldsymbol{W}_1^{(2)}f_1\left(\boldsymbol{W}_1^{(1)}\boldsymbol{x}\right) + \boldsymbol{P}_1\boldsymbol{x}
$$

where $\boldsymbol{W}_1^{(2)}$ is a (10 × 10) parameter matrix that enables more effective linear combinations of the activated inputs. It adds predictive capacity. The number of rows in $\boldsymbol{W}_1^{(1)}$ must be the same as the number of columns in $\boldsymbol{W}_1^{(2)}$. These features are sometimes called the _bottleneck units_. The number of rows in $\boldsymbol{W}_1^{(2)}$ represents the number of hidden units given to the _next_ layer. The matrix need not be square; it may be advantageous to tune it to have more or fewer dimensions than the bottleneck. 

The next two layers also have projection matrices that are (3 × 10) and (10 × 3) for the bottleneck pattern. The equation for future blocks is: 

$$
\boldsymbol{H}_j = \boldsymbol{W}_j^{(2)}f_j\left(\boldsymbol{W}_j^{(1)}\boldsymbol{H}_{j-1}\right) + \boldsymbol{P}_j\boldsymbol{H}_{j-1} 
$$
Residual connections are often used with batch normalization. For tabular data, we would apply the BatchNorm operator $b(\cdot)$ to the embeddings within the activation, such as: 

$$
\boldsymbol{H}_1 = \boldsymbol{W}_1^{(2)}f_1\left(b(\boldsymbol{W}_1^{(1)}\boldsymbol{x})\right) + \boldsymbol{P}_1\boldsymbol{x}
$$

It is difficult to say exactly why residual connections make it easier to train these models. This is partly because residual connections are almost always used in conjunction with batch normalization, as they were in @he2016deep. 

@li2018visualizing makes a compelling case that using residual connections greatly smooths the loss surface and eliminates many jagged regions where gradient descent can get stuck. Additionally, @veit2016residual shows that ResNet models can be rewritten as an additive structure of submodules that are also networks, but many of them are much shallower. They showed that the model might operate more like an ensemble of simpler models and that some of these modules can be ignored without a detrimental effect on performance. 

Let's move on to discuss a technique that has had a phenomenal impact on the quality of neural networks: attention. 

### Attention Mechanisms {#sec-cls-attention}

```{r}
#| label: col-att-start
#| include: false

load("../RData/col_att_demo.RData")
location_df <- col_att_grid |> slice(1:3) |> mutate(label = paste(1:3))
col_att_grid <- col_att_grid |> slice(-(1:3))
```

Attention is a technique that can help a neural network focus on relevant parts of the model to make better individual predictions. It computes a relevance score (an "attention weight") for each input based on its context and use it to modify the embeddings. We'll start by discussing the conceptual idea of what attention does, then segue to how we can use it in tabular models^[Many introductions to attention focus on how it can be used in language models and applications such as sentence completion or translation. See @vaswani2017attention or Section 12 of @bishop2023deep for this type of introduction.]. 

Attention weights are typically created using a query/key/value (QKV) system. This is often described as similar to how a database works. For example, if we had a database of information on published books, we could use it to find our next book to read based on our interests. Suppose the database consisted of a collection of keywords, such as `nonfiction`, `romance`, `history`, and so on, that are used to characterize the book's type and content. The totality of these variables corresponds to the database's _keys_. 

Our _query_ reflects what we are currently interested in. For example, I might make the query:

> I'm interested in a `fictional` `murder-mystery` set in an `isolated location` in `French Canada` with `quirky` characters and a `duck`. 

A database would `FILTER` on these values to find any rows that match the query. Hopefully, the query will return some _values_ that, in this case, correspond to names for relevant books.   

The mathematics of attention mechanisms differ in their tactics, but they pursue the same strategy as the database example above.  Attention computes a quantitative similarity score for the data, usually based on dot products, which are scaled to [0, 1]. This is more of a fuzzy-matching system than an exact match would produce with a SQL `FILTER`. 

Attention weights can be computed in different ways, but we'll focus on the most prevalent approach: *dot-product attention*. We'll represent a query containing $p$ features as the vector $\boldsymbol{q}$, and a database key as $\boldsymbol{k}$, also with $p$ entries. The dot-product attention weight for these values uses function $\mathbb{a}(\boldsymbol{q}, \boldsymbol{k}) = \exp\left(\boldsymbol{q}'\boldsymbol{k}\right)$ as its kernel. From @sec-dot-prod we know that this dot product is non-negative, but generally not much more. 

Most methods use _scaled attention_, which uses the softmax function to compute the attention weights:

$$
\mathbb{a}(\boldsymbol{q}, \boldsymbol{k}_i) = \frac{\exp(\boldsymbol{q}'\boldsymbol{k}_i / \sqrt{p})}{\sum_{i'}\exp(\boldsymbol{q}'\boldsymbol{k}_{i'} / \sqrt{p})},
$$ {#eq-scaled-attention}

where $p$ is the size of the keys and queries. The square-root terms in the divisors are there to improve numerical computations in cases where $p$ is very large; they stabilize the resulting values. 

The function $\mathbb{a}(\boldsymbol{q}, \boldsymbol{k}_i)$ bounds the attention weights to be in [0, 1]. Let's denote a function $\mathbb{S}(\boldsymbol{X})$ that performs row-wise softmax on a matrix $\boldsymbol{X}$ and produces a matrix of the same size whose rows sum to one.

We'll denote original values as the vector $\boldsymbol{v}$. The values could be any characteristic of our data, and we will assume that these are numbers. For our book example, values could include quantities such as the number of pages, the publisher's recommended price, etc. Suppose we are interested in the latter. Let's consider a query that affects a specific key and value pairing to produce an _attended value_: 

$$
\mathbb{A}(\boldsymbol{q}, \boldsymbol{k}_i, \boldsymbol{v}_i) =  \sum_i \mathbb{a}(\boldsymbol{q}, \boldsymbol{k}_i) \boldsymbol{v}_i
$$ {#eq-attention-values}

Our "database" would contain many entries, say $i = 1, \ldots n$, so we would have $n$ **attended values**. For our French Canadian murder-mystery query, this would produce a weighted average of the prices of books with non-zero similarities. The weighting of multiple values can be thought of as _soft attention_, as it blends them with a weighted mean.

::: {.important-box}
In this database analogy, we have a potentially vast set of information (keys and values), and to do anything useful with it, we need a focusing mechanism that aligns with our goals. Our query does just that: it narrows the scope of the data so that our computations are only influenced by the relevant items in the database. In other words, the query provides **context** for our calculations. The literature often refers to the attended values as a "recontextualization of data."
:::

In essence, we are updating our data on a per-query basis.  A different query would generate different weights. As we'll see shortly, attention weights focus the computations on what is relevant for each individual query. 

How is attention incorporated into neural networks? The attention mechanism takes incoming embeddings and produces a set of improved embeddings in a supervised manner. This means that our attention weights need to be learnable parameters so that gradient descent can optimize them. 

Attention is usually followed by one or more layers of traditional MLP-type hidden units that take the preceding embedding and make it more predictive of the outcome. Many neural networks typically use _self-attention_ where $\boldsymbol{X}$ is the query, key, and value elements. 

::: {.mathy-box}
There is a lot of linear algebra on the horizon. In the next section, we'll eventually bring things back down to Earth and demonstrate attention with a small-scale example. 
:::

To be more formal, let's describe the computation for a _single sample_. It enters the attention calculations as $p$ columns, that is, $p$ scalar values (across the batch, an $n \times p$ matrix). These might be the original $p$ predictors, or the hidden units from an earlier layer, depending on the situation.

Because self-attention needs a vector per column to form dot products, the block first embeds each scalar column into an $m$-dimensional vector, turning one sample's $p$ scalars into a ($p \times m$) matrix $\boldsymbol{X}$. Everything below operates on this per-sample matrix; the $n$ samples are each processed the same way, independently^[In code, these per-sample ($p \times m$) matrices are computed together for speed, but the algebra is identical to handling them one at a time.]. The output of the attention block will be a re-mixed set of embeddings $\boldsymbol{Z}$ of the same shape, ($p \times m$). 

Why does self-attention use $\boldsymbol{X}$ for the query, keys, and values? The idea is that, at some point in the network's architecture, we have the current set of embeddings (e.g., the computed hidden units). These are fixed in the sense that their parameters are the same across all samples predicted by the model. With attention, each sample's representation is informed by its own internal feature context. The only type of data used here is the embeddings; the outcome or any other data are not (directly) used. Attention will refocus our embeddings based on the context of each sample^[A shorter answer to the original question is that the individual embeddings are being compared to, and updated by, themselves.]. 

The attention block estimates three sets of weight matrices, all of which are $m \times m$, that project each column's embedding into a query, key, and value: 

$$
\begin{align}
\boldsymbol{Q} &= \boldsymbol{X}\boldsymbol{W}^{q} \notag \\
\boldsymbol{K} &= \boldsymbol{X}\boldsymbol{W}^{k} \notag \\
\boldsymbol{V} &= \boldsymbol{X}\boldsymbol{W}^{v} \notag
\end{align}
$$ {#eq-self-attention}

$\boldsymbol{Q}$, $\boldsymbol{K}$ and $\boldsymbol{V}$ are all $(p \times m)(m \times m) = p \times m$. The $\boldsymbol{W}^*$ matrices are _shared_ across the $p$ columns (the same $\boldsymbol{W}^q$ is applied to every column's embedding) and are the supervised weights optimized during gradient descent. 

To create the attended embeddings: 

$$
\boldsymbol{Z} = \mathbb{A}(\boldsymbol{Q},\boldsymbol{K},\boldsymbol{V})= \mathbb{S}(\boldsymbol{Q}\boldsymbol{K}')\boldsymbol{V} = \boldsymbol{A}\boldsymbol{V}
$$ {#eq-attended-embeddings}

The score matrix $\boldsymbol{Q}\boldsymbol{K}'$ is $(p \times m)(m \times p) = p \times p$: one similarity for every ordered pair of columns. After the row-wise softmax and the product with $\boldsymbol{V}$, the attended embeddings are $(p \times p)(p \times m) = p \times m$. The number of samples $n$ never enters these dimensions; there is simply one ($p \times p$) attention matrix $\boldsymbol{A}$ per sample. This is exactly what allows attention to adapt to each sample: each sample gets its own matrix of column-to-column weights. 

Again, the three $\boldsymbol{W}^*$ matrices are _learned_ during training and, because they are estimated, attention embeddings are indirectly supervised. The target data do _not_ affect the computation's attention during a forward pass; it isn't part of any of the equations above. The target influences the parameters during the backpropagation phase of training (i.e., when the gradients are updated). 

The part of the network that encompasses these calculations is called _an attention head_.  Attention for tabular data primarily occurs in two ways: column attention and row attention. The derivation above, where the $p$ columns attend to one another, is column attention. Row attention instead treats the $n$ samples as the units that attend to one another, so its attention matrix is ($n \times n$). Let's talk about each in turn. 

## Chapter References {.unnumbered}


