---
knitr:
  opts_chunk:
    cache.path: "../_cache/cls-nonlinear/"
---

# Complex Nonlinear Boundaries {#sec-cls-nonlinear}

```{r}
#| label: cls-nonlinear-setup
#| include: false

source("../R/_common.R")

# ------------------------------------------------------------------------------

library(tidymodels)
library(torch)
library(mirai)
library(spatialsample)
library(brulee)
library(bestNormalize)
library(patchwork)

# ------------------------------------------------------------------------------
# set options

tidymodels_prefer()
set_options()
theme_set(theme_bw())
daemons(parallel::detectCores())

# ------------------------------------------------------------------------------
# load data 

load("../RData/forested_data.RData")
load("../RData/three_optimizers.RData")
```

## Nonlinear Discriminants  {#sec-nonlinear-da}

### Quadratic and Regularized Approaches {#sec-qda-rda}

### Mixture Discriminants {#sec-mda}

### Flexible Discriminants {#sec-fda}

### Naive Bayes {#sec-naive-bayes}

## Neural networks via Multilayer Perceptrons {#sec-nnet}

Neural networks are complex models that are comprised of multiple hierarchies of _latent variables_ that represent different motifs in the data. The model’s architecture consists of a series of _layers_, each containing a set of units. Neural networks begin with the input layer that consists of units representing our predictors. After this are one or more _hidden layers_, each containing a preset number of hidden units. These are artificial features that are influenced by the layer before.  For example, if there were four predictors and the first hidden layer had two hidden units, each hidden unit is defined by an equation that uses each of the predictors. In the broadest sense, it is similar to how, in PCA, each principal component is a linear combination of the predictors. However, the similarities between neural networks and PCA end there. These connections between units on different layers continue as layers are added. Eventually, the last layer is the _output layer_. For classification, this usually has as many units as there are class levels in the outcome. @fig-nnet-cls shows a general architectural diagram for a basic model with a single hidden layer. 

```{r}
#| label: fig-nnet-cls
#| echo: false
#| out-width: 40%
#| fig-width: 4
#| fig-height: 5
#| fig-cap: Architecture of a simple one layer neural network.
knitr::include_graphics("../premade/mlp.svg")
```

Traditionally, the connections that come into a hidden unit are combined using a linear predictor: each incoming unit has a corresponding slope, called a "weight" in the neural network literature. The linear combinations commonly have an intercept term (called a "bias"). These linear predictors are embedded into a nonlinear "activation" function ($\sigma_1()$) to enable complex patterns. An example of an activation function would be a sigmoidal function much like the logit relationship used in @eqn-x. 

::: {.note-box}
For most of this section, we’ll be concerned with multilayer perceptrons (MLPs), a subset of the broader class of neural networks. MLPs are feed-forward models, meaning each layer only connects to the subsequent layer. 

As a counterexample, residual networks are connected to previous layers (a.k.a. skip layers); therefore, they are not multilayer perceptrons. 
:::

Here’s a trivial example of an MLP with three features, two outcome classes, and a single layer with four hidden units. If $\sigma_1()$ is our activation function for this layer, the first hidden unit is 

$$
H_{i1} = \sigma(\beta_{01} + \beta_{11}x_{i1} + \ldots + \beta_{41}x_{i4})
$$

```{r}
#| label: nnet-start
#| echo: false
#| out-width: 80%
#| fig-width: 4
#| fig-height: 5
knitr::include_graphics("../premade/nnet_start.svg")
```



The last layer’s unit that corresponds to the first class is computed via

$$
Y_{i1} = \gamma_0 + \gamma_{11}H_{i1} + \ldots + \gamma_{41}H_{i4}
$$

The result is output units $Y$ that are not in proper units to use for classification. To remedy this, the raw outputs are normalized to be probability-like via 

$$
\pi_{ik} = \frac{\exp(Y_{ik})}{\sum\limits_{k=1}^C \exp(Y_{ik})}
$$


```{r}
#| label: nnet-end
#| echo: false
#| out-width: 80%
#| fig-width: 4
#| fig-height: 5
knitr::include_graphics("../premade/nnet_end.svg")
```



look out for exponentiating large numbers

TODO lack of constraints; normalization of inputs

The number of layers, the number of hidden units within a layer, and the type of activation function(s) to use are primary tuning parameters for the MLP structure.  

TODO asis, no preprocessing, imputation, etc is done by the network


TODO advice here is about smaller data sets for tabular mlps; not universally applicable (and most of what is written is erroneously thought to be universally applicable)

In this section, we’ll describe the important elements of multilayer perceptrons, such as the different types of activation functions. We’ll also spend a good deal of time discussing the different ways to train MLPs. There are also discussions on modern deep learning models that are specifically designed for tabular data. 

### Activation Functions

```{r}
#| label: act-grid
#| include: false
lp_grid <- seq(-3, 3, length.out = 100)

act_functions <- 
  bind_rows(
    tibble(input = lp_grid,
           output = binomial()$linkinv(lp_grid),
           activation = "logistic"),
    tibble(input = lp_grid,
           output = tanh(lp_grid),
           activation = "tanh"),
    tibble(input = lp_grid,
           output = as.numeric(nnf_relu(lp_grid)),
           activation = "relu"),
    tibble(input = lp_grid,
           output = as.numeric(nnf_elu(lp_grid)),
           activation = "elu"),
    tibble(input = lp_grid,
           output = lp_grid - tanh(lp_grid),
           activation = "tanhshrink"),
    tibble(input = lp_grid,
           output = as.numeric(nnf_softshrink(lp_grid)),
           activation = "softshrink"),
    tibble(input = lp_grid,
           output = as.numeric(nnf_gelu(lp_grid)),
           activation = "gelu"),
    tibble(input = lp_grid,
           output = as.numeric(nnf_silu(lp_grid)),
           activation = "silu"),
    tibble(input = lp_grid,
           output = as.numeric(nn_log_sigmoid()(lp_grid)),
           activation = "log(sigmoid)"),
    tibble(input = lp_grid,
           output = as.numeric(nn_softplus()(lp_grid)),
           activation = "softplus")
  )
```  

Historically, the most commonly used nonlinear activation functions were sigmoidal in nature. This means that the linear combination of the elements of the previous layer was converted into a "squashed" format, usually between zero and one. @fig-sigmoid-activations shows several such functions. The logistic (eq TODO) function and hyperbolic tangent are fairly similar, and were typical defaults. The tanshrink function diminishes the impact of inputs that are in a neighborhood around zero. For large input values, the output is linear in the output (with an offset of $\pm 1$). 

```{r}
#| label: fig-sigmoid-activations
#| echo: false
#| out-width: 85%
#| fig-width: 8.25
#| fig-height: 2
#| fig-cap: "Different sigmoidal activation functions."

sig_lvl <- c("logistic", "tanh", "tanhshrink", "log(sigmoid)")

act_functions |>  
  filter(grepl("(tan)|(sig)|(logistic)", activation)) |> 
  mutate(activation = factor(activation, levels = sig_lvl)) |> 
  ggplot(aes(input, output)) + 
  geom_line() +
  facet_wrap(~ activation, scale = "free_y", nrow = 1) + 
  theme_bw()
```

In modern neural networks, the rectified linear unit (ReLU) is the default activation function. As shown in @fig-relu-activations, the function resembles the hinge functions used by MARS; to the left-hand side of zero, the output is zero, and then it is linear otherwise. Much like MARS, this has the effect of isolating the effect of some region of the predictor space. However, there are a few key differences. MARS hinge functions affect a single predictor at a time, while neural networks’ activation functions act on the linear combination of parameters. Also, neural networks with multiple layers recursively nest actions of linear combinations inside the prior layers (and so on). The mathematical function is very similar, but the application of this function has a more comprehensive impact on the model. 

There are various activation functions that are modified versions of ReLU that emulate the same form but have continuous derivatives. These are also shown in @fig-relu-activations, including softplus [@dugas2000incorporating;@glorot2011deep], gelu [@hendrycks2016gaussian], silu [@elfwing2018sigmoid], and elu [@clevert2015fast]. 

```{r}
#| label: fig-relu-activations
#| echo: false
#| out-width: 90%
#| fig-width: 8
#| fig-height: 2
#| fig-cap: "Different ReLU-based activation functions."

relu_lvl <- c("relu", "softplus", "gelu", "silu", "elu")

act_functions |>  
  filter(grepl("(l[uU]$)|(softplus)", activation)) |> 
  mutate(activation = factor(activation, levels = relu_lvl)) |> 
  ggplot(aes(input, output)) + 
  geom_hline(yintercept = 0, lty = 3) + 
  geom_line() +
  facet_wrap(~ activation, nrow = 1) + 
  theme_bw()
```

```{r}
#| label: nnet-ex-comps
#| include: false
#| cache: true
two_class_ex <- 
  parabolic |> 
  set_names(c("A", "B", "Class")) 

two_class_ex <-   
  recipe(Class ~ ., data = two_class_ex) |> 
  step_orderNorm(A, B) |> 
  prep() |> 
  bake(new_data = NULL)

x <- seq(-4, 4, length.out = 100)
grid <- crossing(A = x, B = x)

set.seed(123)
fit_1 <- brulee_mlp(
  Class ~ A + B,
  data = two_class_ex,
  hidden_units = 3,
  penalty = 0,
  activation = "relu",
  verbose = TRUE
)

coefs <- coef(fit_1)

grid_pred <- 
  fit_1 |> 
  predict(grid, type = "prob") |> 
  bind_cols(grid)

get_elements <- function(x, model) {
  xtbl <- x
  x <- as.matrix(x)
  lp_01 <- x %*% t(coefs$model.0.weight) + coefs$model.0.bias
  colnames(lp_01) <- paste0("lp_", 1:ncol(lp_01))
  layer_01 <- as.matrix(torch::nnf_relu(lp_01))
  colnames(layer_01) <- paste0("relu_", 1:ncol(layer_01))
  lp_12 <- layer_01 %*% t(coefs$model.2.weight) + coefs$model.2.bias
  colnames(lp_12) <- paste0("output_", 1:ncol(lp_12))
  prob <- exp(lp_12) / sum(exp(lp_12))
  bind_cols(xtbl, as_tibble(lp_01), as_tibble(layer_01), as_tibble(lp_12))
}

components <- NULL
for (i in 1:nrow(grid)) {
  components <- bind_rows(components, get_elements(grid[i,], fit_1))
}
```

Let’s use an example data set with a simple MLP to demonstrate how ReLU activation works. @fig-nnet-ex shows the end results of a model with a single layer containing three hidden units with ReLU activation. The model converged in `r fit_1$best_epoch` iterations and does a good job discriminating between the two classes. 

```{r}
#| label: fig-nnet-ex
#| echo: false
#| out-width: 60%
#| fig-width: 4.4
#| fig-height: 4.4
#| warning: false
#| fig-cap: "An example data set with two predictors and two classes. The black line is the class boundary from an MLP model using three hidden units that use ReLU activation."

  grid_pred |> 
  ggplot(aes(A, B)) +
  geom_point(data = two_class_ex, aes(col = Class), alpha = 1/3) +
  geom_contour(aes(z = .pred_Class1), breaks = 0.5, col = "black") +
  coord_fixed() +
  scale_color_manual(values = c("#007FFFFF", "#FF7F00FF"))
```

With two predictors and three hidden units, the first set of coefficients for the model creates three sets of linear predictors (i.e., an intercept and two slopes) for each of the three hidden units. The predictors were normalized to have the same mean and variance (zero and one, respectively). The three equations produced by training were: 

```{r}
#| label: nnet-ex-first
#| echo: false
#| results: "asis"
wt_1 <- format(coefs$model.0.weight, digits = 1)
wt_1 <- gsub("^ ", "+", wt_1)
wt_1[,1] <- paste0(wt_1[,1], "\\:A_i")
wt_1[,2] <- paste0(wt_1[,2], "\\:B_i")
lp_1 <- 
  cbind(
    format(coefs$model.0.bias, digits = 2),
    wt_1
  )
lp_1 <- apply(lp_1, 1, function(x) paste0(x, collapse = ""))
lp_1 <- paste0("H_{i", 1:3, "}&=", lp_1, " \\notag")

cat("$$")
cat("\\begin{align}")
cat(paste0(lp_1, collapse = " \\\\ \n"))
cat("\\end{align}")
cat("$$")
```

The intercept estimate for the first hidden unit is much larger than the other two. That unit also has a large positive slope for predictor A. The second hidden unit emphasizes predictor B, while $H_{i3}$ has strong positive effects for both predictors. 

The top of @fig-nnet-ex-layer-1 shows a heatmap of how these three artificial features change over the two-dimensional predictor space. The black line segments show the contour corresponding to zero values. The first hidden unit has larger values on the right-hand side of the predictor space, with a slightly diagonal contour for zero. The pattern shown for the second hidden unit is nearly the opposite of the first, although the zero contour cuts more through the center of the space. Finally, the third unit is mostly driven by predictor B with larger values in the upper right region. 

```{r}
#| label: fig-nnet-ex-layer-1
#| echo: false
#| out-width: 80%
#| fig-width: 7
#| fig-height: 5
#| warning: false
#| fig-cap: "Heatmaps of the three linear predictors and their ReLU values across the predictor space. The black line emphasizes the location where the linear predictor is zero."

lp_layer <- 
  components |> 
  select(-starts_with("relu_"), -starts_with("output")) |> 
  pivot_longer(cols = c(starts_with("lp_"))) |> 
  mutate(name = gsub("lp_", "hidden unit ", name)) |> 
  ggplot(aes(A, B)) + 
  geom_tile(aes(fill = value), show.legend = TRUE) + 
  geom_contour(aes(z = value), breaks = 0, col = "black") +
  facet_wrap(~ name, nrow = 1) + 
  scale_fill_gradient2(high = "#5D74A5FF") + 
  # coord_fixed() + 
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  ) + 
  labs(title = "Linear Predictor")


relu_layer <- 
  components |> 
  select(-starts_with("lp"), -starts_with("output")) |> 
  pivot_longer(cols = c(starts_with("relu_"))) |> 
  mutate(name = gsub("relu_", "hidden unit ", name)) |> 
  ggplot(aes(A, B)) + 
  geom_tile(aes(fill = value), show.legend = TRUE) + 
  geom_contour(aes(z = value), breaks = 0, col = "5AAE61FF") +
  facet_wrap(~ name, nrow = 1) + 
  scale_fill_gradient2(high = "#5D74A5FF") + 
  # coord_fixed() + 
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  ) + 
  labs(title = "ReLU", y = NULL) 

lp_layer / relu_layer
```

The bottom side of the figure shows the effect of the ReLU function; the values less than zero are set to zero. At this point, $H_{i1}$ and $H_{i3}$ have the strongest effect on the model, but in somewhat overlapping regions. $H_{i1}$ is mostly active for smaller values of predictor A. 

However, the model estimates additional parameters that will merge these regions with weights that are most helpful for predicting the data. Those equations were estimated to be: 

```{r}
#| label: nnet-ex-last
#| echo: false
#| results: "asis"
wt_2 <- format(coefs$model.2.weight, digits = 1)
wt_2 <- gsub("^ ", "+", wt_2)
wt_2[,1] <- paste0(wt_2[,1], "H_{i1}")
wt_2[,2] <- paste0(wt_2[,2], "H_{i2}")
wt_2[,3] <- paste0(wt_2[,3], "H_{i3}")
lp_2 <- 
  cbind(
    format(coefs$model.2.bias, digits = 2),
    wt_2
  )
lp_2 <- apply(lp_2, 1, function(x) paste0(x, collapse = ""))
lp_2 <- paste0("Y_{i", 1:2, "}&=", lp_2, " \\notag")

cat("\\begin{align}")
cat(paste0(lp_2, collapse = " \\\\ \n"))
cat("\\end{align}")
```

Note that the coefficients’ signs are opposite for each of the outcome units. @fig-nnet-ex-layer-2 illustrates the transition from the hidden layer to the outcome layer. The weighted sum of the three variables at the top produces the regions of strong positive (or negative) activation for the two classes. Once again, the black curve shows where the zero contour is located. 

```{r}
#| label: fig-nnet-ex-layer-2
#| echo: false
#| out-width: 80%
#| fig-width: 7
#| fig-height: 5
#| warning: false
#| fig-cap: "Heatmaps of the three ReLU units and their output units across the predictor space."

output_layer <- 
  components |>
  select(-starts_with("lp"), -starts_with("relu_")) |>
  pivot_longer(cols = c(starts_with("output"))) |>
  ggplot(aes(A, B)) +
  geom_tile(aes(fill = value)) +
  geom_contour(
    # data = grid_pred,
    aes(z = value),
    breaks = 0,
    col = "black"
  ) +
  facet_wrap(~ name, nrow = 1) + 
  scale_fill_gradient2(high = "#420F75FF", low = "#EAC541FF") + 
  coord_fixed() + 
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  )  + 
  labs(title = "Output", y = NULL) 

relu_layer / output_layer
```

The two panels show nearly opposite patterns, which is not coincidental. The model converges to this structure so that the output units would favor one class or the other. The legend on the bottom right is clearly in the wrong units; we’d like them to resemble probabilities. As previously mentioned, the softmax function coerces these values to be on [0, 1] and to also sum to one. This produces the class boundary seen in @fig-nnet-ex. 

As previously stated, ReLU is not a smooth function, so this demonstration will be somewhat different from the other activation functions; they generally don’t completely isolate regions of the predictor space. However, the general flow of the network is still the same. 

### Training and Optimization

It can be challenging to train neural network classifiers. Their model structure is composed of multiple nested nonlinear functions that are not smooth, continuous functions for activations such as ReLU. The loss function, commonly cross-entropy, is unlikely to be convex, especially as model complexity increases. As a consequence, we have no guarantees that a gradient-based optimizer will converge. There is also the possibility of local minima that might trap the training process, saddle points that could lead the search in the wrong direction, and regions of vanishing gradients. These issues will be discussed below. 

Another possible factor is that many deep learning models are trained on vast amounts of data. This in itself prevents constraints on how much data can be held in memory at once, and other logistical issues. This may not be an issue for the average application that uses tabular data, but the problem still exists. 

Before proceeding, recall @sec-gradient-opt where gradient descent was introduced. There are two commonly used classes of gradient-based optimization methods:  _first-order techniques_, where the updating equation is: 

$$
\boldsymbol{\theta}_{i+1} = \boldsymbol{\theta}_i - \alpha\:g(\boldsymbol{\theta}_i)
$$, 

and _second-order methods_: 

$$\boldsymbol{\theta}_{i+1} = \boldsymbol{\theta}_i - \alpha H(\boldsymbol{\theta_i})^{-1}\:g(\boldsymbol{\theta}_i) 
$$,

where $\boldsymbol{\theta}$ are the unknown parameters, $\alpha$ is the learning rate, $g(\boldsymbol{\theta}_i)$ is the gradient vector of first derivatives, and $H(\boldsymbol{\theta_i})$ is the Hessian matrix of second derivatives. The Hessian measures the curvature of the loss function at the current iteration. Pre-multiplying by its inverse rescales the gradient vector, generally making it more effective. We'll discuss them both in @sec-gradient-descent and @sec-newtons-method below. In most cases, second-order methods are preferred since they are more accurate. However, first-order search is more commonly used because it is faster to compute and can be much more memory efficient^[Making it infeasible to compute efficiently when the training set is very large.]

To demonstrate, an MLP model was used to model the forestation data that contained one layer of 30 ReLU units and an additional layer of 5 ReLU units. An L<sub>2</sub> penalty of 0.15 was used to regularize the cross-entropy loss function. A maximum of 100 epochs was specified, but training would stop after five bad epochs of validation loss. An initial learning rate of $\alpha = 0.1$ was used, but a variable, step-wise rate schedule was used to modulate the rate across epochs (described below). 

The model was first trained using gradient descent. For illustration, training was executed five times using different initial parameter values. The leftmost panel in @fig-three-optimizers shows the results. In each case, the loss function had an initial middling reduction. In each case, the model used all 100 epochs. This approach was not very successful. There was not much of a reduction in the loss function, and the terminal phase had a minuscule decrease with no real improvement. 

```{r}
#| label: fig-three-optimizers
#| echo: false
#| out-width: 80%
#| fig-width: 6
#| fig-height: 2.8
#| warning: false
#| fig-cap: "The validation loss function for the same MLP model for three optimzation strategies. The lines on each panel represents runs initialized with different random numbers."

three_optimizers |> 
  ggplot(aes(epoch, loss_norm, col = seed, group = seed)) +
  geom_line(show.legend = FALSE,
            linewidth = 3 / 4,
            alpha = 1 / 2) +
  facet_wrap(~ method, scale = "free_x") +
  scale_x_continuous(breaks = pretty_breaks()) + 
  labs(x = "Epochs", y = "Cross-Entropy (Validation)")
```

As a second approach, an approximation to Newton’s Method called the Limited-memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS) method. This uses an approximation to the Hessian matrix. In this case, there was an immediate and enormous reduction in loss in the first epoch. In most cases, the next few epochs resulted in further reductions. Some of the training runs quickly converged but one required 14 eopchs. Compared to basic gradient descent, L-BFGS produced a much more effective model (based on the loss). The median time for optimization was `r round(three_optimizer_times$median[three_optimizer_times$method == "L-BFGS"], 1)`s, which was somewhat faster than the median time for gradient descent (`r round(three_optimizer_times$median[three_optimizer_times$method == "GD"], 1)`s). _However_, we should consider the time per epoch since GD used all allowed epochs. L-BFGS required `r round(three_optimizer_times$per_epoch[three_optimizer_times$method == "L-BFGS"], 3)`s per epoch while gradient descent required `r round(three_optimizer_times$per_epoch[three_optimizer_times$method == "GD"], 3)`s

As-is, gradient descent might be deemed ineffectual, and a second-order approach like L-BFGS might be the best approach. However, the plain version of gradient descent is almost never used. Instead, different variations of _stochastic_ gradient descent (SGD) are the norm. The rightmost panel in @fig-three-optimizers shows the results when the _Nesterov momentum_ term was used. This initially computes a standard update to the parameters, but then approximates the gradient at this position to compute a second prospective step. The actual update that is used is the lookahead position. 

The stochastic aspect of the search used randomly allocated _batches_ of 64 training set samples. The gradient search is updated by processing each of these batches until the entire trained set has been used, which signifies the end of that epoch. As a result, for this training set size, about 76  gradient updates are executed before the end of each epoch. 

Despite making many potentially low-quality gradient steps, SGD almost always produces better results than one gradient step using all of the training set. This is the case for our model with two layers; the results are just as good as the L-BFGS search. SGD was much slower overall, using `r  round(three_optimizer_times$median[three_optimizer_times$method == "SGD with momentum"], 1)`s but its per epoch time was between the other two methods (`r round(three_optimizer_times$per_epoch[three_optimizer_times$method == "SGD with momentum"], 3)`s).

The main takeaways of this exercise are that training these models can be difficult and that the type of parameter optimization can substantially impact model quality. 

<hr>

Saddle points occur when the gradient is zero, but there are regions surrounding this location that are increasing and others that are decreasing. 

This situation can result in epochs where the search stagnates and shows barely incremental progress. As the search approaches the saddle point, the gradient becomes very close to zero, and first-order gradient searches have very small updates.

#### Initialization

Network parameter initialization can significantly impact training success. Given the size and complexity of these models, optimal methodologies for setting initial values are nearly impossible to derive. As such, parameters are initialized using random numbers. There are a few approaches to do this. @lecun2002efficient initialization uses a uniform distribution centered around zero with limits defined by the number of connections feeding into the current node (known as the "fan in" and is denoted as $p_{in}$). They used $U(\pm \sqrt{2/p_{in}})$ as the distribution. There have been a few variations of this: 

 - Glorot/Xavier initialization [@glorot2010understanding]: $U(\pm \sqrt{6/(p_{in}+p_{out})})$ 
 - He/Kaiming initialization [@he2015delving]: $N(0, \sqrt{2 / p_{in}})$.

One reason that initialization is important is to help mitigate numerical overflow errors. For example, the softmax transformation shown above exponentiates the raw output units (and sums them). If the output units have relatively large values, exponentiating and summing them can exceed allowable ranges. For example, the output units shown in @fig-nnet-ex-layer-2 have a wide enough range that can be as large as `r max(round(abs(components$output_1)), 0)`. 

#### Learning Rates

The learning rate ($\alpha$) can have a profound effect on model quality, perhaps more than the structural tuning parameters that define complexity (e.g., number of hidden units). A large learning range will help converge towards effective parameter estimates, but once there, it can cause the model to consistently overshoot the best. One approach to modulating the rate is to use a scheduler [@lu2022gradient,chapt 4]. These are functions/algorithms that will change the rate (often to decrease it) over epochs. For example, to monotonically decrease the rate, two options are:

- inverse decay: $f(\alpha, \delta, i) = \alpha_0/(1 + \delta i)$ 
- exponential decay: $f(\alpha, \delta, i) = \alpha_0\exp(-\delta i)$

where $i$ is the epoch number and $\delta$ is a decay value. These two functions gradually determine the rate so that, when near an optimum, it is slow enough not to exceed the optimal settings. Another approach is a step function that decreases the rate but holds it constant for a range of epochs. Those ranges can be predefined or can be set dynamically so that the rate decreases once the loss function is reduced. The function for the former is: 

- step decay: $f(\alpha, \zeta, i) = \alpha_0 \zeta^{\lfloor i/r\rfloor}$ 

where $\zeta$ is the reduction value, $r$ is the step length, and $\lfloor x\rfloor$ is the floor function. 

Finally, there are schedulers that modulate the rate up and down. A cyclic schedule [@smith2017cyclical] initially starts at a maximum value, decreases to a minimum value, and then cyclically increases again. For some range $r$ and minimum rate $\alpha_0$, an epoch's location within the series is $cycle = \lfloor 1 + (i / (2r) )\rfloor$ and the rate is

- cyclic: $f(\alpha_0, \alpha, r) = \alpha_0 + ( \alpha - \alpha_0 ) \max(0, 1 - x)$

where $x = |(i/r) - (2\:cycle) + 1|$. This is a good option when we are not sure how fast or slow the convergence will be. If the rate is too high near an optimum value, it will soon decrease to an appropriate value. Also, if the search is stuck in a local optimum (or a saddle-point), moving back to a larger learning rate might help it escape. 

The type of schedule, and its additional parameter(s), are added to the list of possible values that could be optimized during tuning.

#### Early Stopping {#sec-early-stopping}

#### Batch Learning {#sec-batch-learning}

#### Gradient Descent {#sec-gradient-descent}

#### Newton’s Method {#sec-newtons-method}

### Data Characteristics

### Forestation Modeling

## Special Tabular Network Models  {#sec-tab-net}

### Transformers

### Attention


## K-Nearest Neighbors  {#sec-cls-knn}

### Discriminant Adaptive Nearest Neighbors  {#sec-dann}

## Support Vector Machines {#sec-cls-svm}

## Chapter References {.unnumbered}
