---
knitr:
  opts_chunk:
    cache.path: "../_cache/cls-nonlinear/"
---

# Complex Nonlinear Boundaries {#sec-cls-nonlinear}

```{r}
#| label: cls-nonlinear-setup
#| include: false

source("../R/_common.R")

# ------------------------------------------------------------------------------

library(tidymodels)
library(torch)
library(mirai)
library(spatialsample)
library(brulee)
library(bestNormalize)
library(patchwork)

# ------------------------------------------------------------------------------
# set options

tidymodels_prefer()
set_options()
theme_set(theme_bw())
daemons(parallel::detectCores())

# ------------------------------------------------------------------------------
# load data 

load("../RData/forested_data.RData")
load("../RData/optimizers.RData")

# ------------------------------------------------------------------------------
# Gradient descent helpers

source("../R/setup_2D_GD.R")
```

## Nonlinear Discriminants  {#sec-nonlinear-da}

### Quadratic and Regularized Approaches {#sec-qda-rda}

### Mixture Discriminants {#sec-mda}

### Flexible Discriminants {#sec-fda}

### Naive Bayes {#sec-naive-bayes}

## Neural Networks via Multilayer Perceptrons {#sec-nnet}

Neural networks are complex models that are comprised of multiple hierarchies of _latent variables_ that represent different motifs in the data. The model’s architecture consists of a series of _layers_, each containing a set of units. Neural networks begin with the input layer that consists of units representing our predictors. After this are one or more _hidden layers_, each containing a preset number of hidden units. These are artificial features that are influenced by the layer before.  For example, if there were four predictors and the first hidden layer had two hidden units, each hidden unit is defined by an equation that uses each of the predictors. In the broadest sense, it is similar to how, in PCA, each principal component is a linear combination of the predictors. However, the similarities between neural networks and PCA end there. These connections between units on different layers continue as layers are added. Eventually, the last layer is the _output layer_. For classification, this usually has as many units as there are class levels in the outcome. @fig-nnet-cls shows a general architectural diagram for a basic model with a single hidden layer. 

```{r}
#| label: fig-nnet-cls
#| echo: false
#| out-width: 40%
#| fig-width: 4
#| fig-height: 5
#| fig-cap: Architecture of a simple neural network with a single layer of hidden units.
knitr::include_graphics("../premade/mlp.svg")
```

Traditionally, the connections that come into a hidden unit are combined using a linear predictor: each incoming unit has a corresponding slope, called a "weight" in the neural network literature. The linear combinations commonly have an intercept term (called a "bias"). These linear predictors are embedded into a nonlinear "activation" function ($\sigma$) to enable complex patterns. An example of an activation function would be a sigmoidal function much like the logit relationship used in @eqn-x. 

::: {.note-box}
For most of this section, we’ll be concerned with multilayer perceptrons (MLPs), a subset of the broader class of neural networks. MLPs are feed-forward models, meaning each layer only connects to the subsequent layer. 

As a counterexample, residual networks are connected to previous layers (a.k.a. skip layers); therefore, they are not multilayer perceptrons. 
:::

Here’s a trivial example of an MLP with three features, two outcome classes, and a single layer with four hidden units. If $\sigma_1()$ is our activation function for this layer, the first hidden unit is 

$$
H_{i1} = \sigma(\beta_{01} + \beta_{11}x_{i1} + \ldots + \beta_{41}x_{i4})
$$

```{r}
#| label: nnet-start
#| echo: false
#| out-width: 80%
#| fig-width: 4
#| fig-height: 5
knitr::include_graphics("../premade/nnet_start.svg")
```

The last layer’s unit that corresponds to the first class is computed via
$$
Y_{i1} = \gamma_0 + \gamma_{11}H_{i1} + \ldots + \gamma_{41}H_{i4}
$$

The result is output units $Y$ that are not in proper units to use for classification. To remedy this, the raw outputs are normalized to be probability-like via 

$$
\pi_{ik} = \frac{\exp(Y_{ik})}{\sum\limits_{k=1}^C \exp(Y_{ik})}
$$

```{r}
#| label: nnet-end
#| echo: false
#| out-width: 80%
#| fig-width: 4
#| fig-height: 5
knitr::include_graphics("../premade/nnet_end.svg")
```

look out for exponentiating large numbers

TODO lack of constraints; normalization of inputs

The number of layers, the number of hidden units within a layer, and the type of activation function(s) to use are primary tuning parameters for the MLP structure.  

TODO asis, no preprocessing, imputation, etc is done by the network

TODO advice here is about smaller data sets for tabular mlps; not universally applicable (and most of what is written is erroneously thought to be universally applicable)

In this section, we’ll describe the important elements of multilayer perceptrons, such as the different types of activation functions. We’ll also spend a good deal of time discussing the different ways to train MLPs. There are also discussions on modern deep learning models that are specifically designed for tabular data. 

### Activation Functions {#sec-nnet-activation}

```{r}
#| label: act-grid
#| include: false
lp_grid <- seq(-3, 3, length.out = 100)

act_functions <- 
  bind_rows(
    tibble(input = lp_grid,
           output = binomial()$linkinv(lp_grid),
           activation = "logistic"),
    tibble(input = lp_grid,
           output = tanh(lp_grid),
           activation = "tanh"),
    tibble(input = lp_grid,
           output = as.numeric(nnf_relu(lp_grid)),
           activation = "relu"),
    tibble(input = lp_grid,
           output = as.numeric(nnf_elu(lp_grid)),
           activation = "elu"),
    tibble(input = lp_grid,
           output = lp_grid - tanh(lp_grid),
           activation = "tanhshrink"),
    tibble(input = lp_grid,
           output = as.numeric(nnf_softshrink(lp_grid)),
           activation = "softshrink"),
    tibble(input = lp_grid,
           output = as.numeric(nnf_gelu(lp_grid)),
           activation = "gelu"),
    tibble(input = lp_grid,
           output = as.numeric(nnf_silu(lp_grid)),
           activation = "silu"),
    tibble(input = lp_grid,
           output = as.numeric(nn_log_sigmoid()(lp_grid)),
           activation = "log(sigmoid)"),
    tibble(input = lp_grid,
           output = as.numeric(nn_softplus()(lp_grid)),
           activation = "softplus")
  )
```  

Historically, the most commonly used nonlinear activation functions were sigmoidal in nature. This means that the linear combination of the elements of the previous layer was converted into a "squashed" format, usually between zero and one. @fig-sigmoid-activations shows several such functions. The logistic (eq TODO) function and hyperbolic tangent are fairly similar, and were typical defaults. The tanshrink function diminishes the impact of inputs that are in a neighborhood around zero. For large input values, the output is linear in the output (with an offset of $\pm 1$). 

```{r}
#| label: fig-sigmoid-activations
#| echo: false
#| out-width: 85%
#| fig-width: 8.25
#| fig-height: 2
#| fig-cap: "Different sigmoidal activation functions."

sig_lvl <- c("logistic", "tanh", "tanhshrink", "log(sigmoid)")

act_functions |>  
  filter(grepl("(tan)|(sig)|(logistic)", activation)) |> 
  mutate(activation = factor(activation, levels = sig_lvl)) |> 
  ggplot(aes(input, output)) + 
  geom_line() +
  facet_wrap(~ activation, scale = "free_y", nrow = 1) + 
  theme_bw()
```

In modern neural networks, the rectified linear unit (ReLU) is the default activation function. As shown in @fig-relu-activations, the function resembles the hinge functions used by MARS; to the left-hand side of zero, the output is zero, and then it is linear otherwise. Much like MARS, this has the effect of isolating the effect of some region of the predictor space. However, there are a few key differences. MARS hinge functions affect a single predictor at a time, while neural networks’ activation functions act on the linear combination of parameters. Also, neural networks with multiple layers recursively nest actions of linear combinations inside the prior layers (and so on). The mathematical function is very similar, but the application of this function has a more comprehensive impact on the model. 

There are various activation functions that are modified versions of ReLU that emulate the same form but have continuous derivatives. These are also shown in @fig-relu-activations, including softplus [@dugas2000incorporating;@glorot2011deep], gelu [@hendrycks2016gaussian], silu [@elfwing2018sigmoid], and elu [@clevert2015fast]. 

```{r}
#| label: fig-relu-activations
#| echo: false
#| out-width: 90%
#| fig-width: 8
#| fig-height: 2
#| fig-cap: "Different ReLU-based activation functions."

relu_lvl <- c("relu", "softplus", "gelu", "silu", "elu")

act_functions |>  
  filter(grepl("(l[uU]$)|(softplus)", activation)) |> 
  mutate(activation = factor(activation, levels = relu_lvl)) |> 
  ggplot(aes(input, output)) + 
  geom_hline(yintercept = 0, lty = 3) + 
  geom_line() +
  facet_wrap(~ activation, nrow = 1) + 
  theme_bw()
```

```{r}
#| label: nnet-ex-comps
#| include: false
#| cache: true
two_class_ex <- 
  parabolic |> 
  set_names(c("A", "B", "Class")) 

two_class_ex <-   
  recipe(Class ~ ., data = two_class_ex) |> 
  step_orderNorm(A, B) |> 
  prep() |> 
  bake(new_data = NULL)

x <- seq(-4, 4, length.out = 100)
grid <- crossing(A = x, B = x)

set.seed(123)
fit_1 <- brulee_mlp(
  Class ~ A + B,
  data = two_class_ex,
  hidden_units = 3,
  penalty = 0,
  activation = "relu",
  verbose = TRUE
)

coefs <- coef(fit_1)

grid_pred <- 
  fit_1 |> 
  predict(grid, type = "prob") |> 
  bind_cols(grid)

get_elements <- function(x, model) {
  xtbl <- x
  x <- as.matrix(x)
  lp_01 <- x %*% t(coefs$model.0.weight) + coefs$model.0.bias
  colnames(lp_01) <- paste0("lp_", 1:ncol(lp_01))
  layer_01 <- as.matrix(torch::nnf_relu(lp_01))
  colnames(layer_01) <- paste0("relu_", 1:ncol(layer_01))
  lp_12 <- layer_01 %*% t(coefs$model.2.weight) + coefs$model.2.bias
  colnames(lp_12) <- paste0("output_", 1:ncol(lp_12))
  prob <- exp(lp_12) / sum(exp(lp_12))
  bind_cols(xtbl, as_tibble(lp_01), as_tibble(layer_01), as_tibble(lp_12))
}

components <- NULL
for (i in 1:nrow(grid)) {
  components <- bind_rows(components, get_elements(grid[i,], fit_1))
}
```

Let’s use an example data set with a simple MLP to demonstrate how ReLU activation works. @fig-nnet-ex shows the end results of a model with a single layer containing three hidden units with ReLU activation. The model converged in `r fit_1$best_epoch` iterations and does a good job discriminating between the two classes. 

```{r}
#| label: fig-nnet-ex
#| echo: false
#| out-width: 60%
#| fig-width: 4.4
#| fig-height: 4.4
#| warning: false
#| fig-cap: "An example data set with two predictors and two classes. The black line is the class boundary from an MLP model using three hidden units that use ReLU activation."

  grid_pred |> 
  ggplot(aes(A, B)) +
  geom_point(data = two_class_ex, aes(col = Class), alpha = 1/3) +
  geom_contour(aes(z = .pred_Class1), breaks = 0.5, col = "black") +
  coord_fixed() +
  scale_color_manual(values = c("#007FFFFF", "#FF7F00FF"))
```

With two predictors and three hidden units, the first set of coefficients for the model creates three sets of linear predictors (i.e., an intercept and two slopes) for each of the three hidden units. The predictors were normalized to have the same mean and variance (zero and one, respectively). The three equations produced by training were: 

```{r}
#| label: nnet-ex-first
#| echo: false
#| results: "asis"
wt_1 <- format(coefs$model.0.weight, digits = 1)
wt_1 <- gsub("^ ", "+", wt_1)
wt_1[,1] <- paste0(wt_1[,1], "\\:A_i")
wt_1[,2] <- paste0(wt_1[,2], "\\:B_i")
lp_1 <- 
  cbind(
    format(coefs$model.0.bias, digits = 2),
    wt_1
  )
lp_1 <- apply(lp_1, 1, function(x) paste0(x, collapse = ""))
lp_1 <- paste0("H_{i", 1:3, "}&=", lp_1, " \\notag")

cat("$$")
cat("\\begin{align}")
cat(paste0(lp_1, collapse = " \\\\ \n"))
cat("\\end{align}")
cat("$$")
```

The intercept estimate for the first hidden unit is much larger than the other two. That unit also has a large positive slope for predictor A. The second hidden unit emphasizes predictor B, while $H_{i3}$ has strong positive effects for both predictors. 

The top of @fig-nnet-ex-layer-1 shows a heatmap of how these three artificial features change over the two-dimensional predictor space. The black line segments show the contour corresponding to zero values. The first hidden unit has larger values on the right-hand side of the predictor space, with a slightly diagonal contour for zero. The pattern shown for the second hidden unit is nearly the opposite of the first, although the zero contour cuts more through the center of the space. Finally, the third unit is mostly driven by predictor B with larger values in the upper right region. 

```{r}
#| label: fig-nnet-ex-layer-1
#| echo: false
#| out-width: 80%
#| fig-width: 7
#| fig-height: 5
#| warning: false
#| fig-cap: "Heatmaps of the three linear predictors and their ReLU values across the predictor space. The black line emphasizes the location where the linear predictor is zero."

lp_layer <- 
  components |> 
  select(-starts_with("relu_"), -starts_with("output")) |> 
  pivot_longer(cols = c(starts_with("lp_"))) |> 
  mutate(name = gsub("lp_", "hidden unit ", name)) |> 
  ggplot(aes(A, B)) + 
  geom_tile(aes(fill = value), show.legend = TRUE) + 
  geom_contour(aes(z = value), breaks = 0, col = "black") +
  facet_wrap(~ name, nrow = 1) + 
  scale_fill_gradient2(high = "#5D74A5FF") + 
  # coord_fixed() + 
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  ) + 
  labs(title = "Linear Predictor")

relu_layer <- 
  components |> 
  select(-starts_with("lp"), -starts_with("output")) |> 
  pivot_longer(cols = c(starts_with("relu_"))) |> 
  mutate(name = gsub("relu_", "hidden unit ", name)) |> 
  ggplot(aes(A, B)) + 
  geom_tile(aes(fill = value), show.legend = TRUE) + 
  geom_contour(aes(z = value), breaks = 0, col = "5AAE61FF") +
  facet_wrap(~ name, nrow = 1) + 
  scale_fill_gradient2(high = "#5D74A5FF") + 
  # coord_fixed() + 
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  ) + 
  labs(title = "ReLU", y = NULL) 

lp_layer / relu_layer
```

The bottom side of the figure shows the effect of the ReLU function; the values less than zero are set to zero. At this point, $H_{i1}$ and $H_{i3}$ have the strongest effect on the model, but in somewhat overlapping regions. $H_{i1}$ is mostly active for smaller values of predictor A. 

However, the model estimates additional parameters that will merge these regions with weights that are most helpful for predicting the data. Those equations were estimated to be: 

```{r}
#| label: nnet-ex-last
#| echo: false
#| results: "asis"
wt_2 <- format(coefs$model.2.weight, digits = 1)
wt_2 <- gsub("^ ", "+", wt_2)
wt_2[,1] <- paste0(wt_2[,1], "H_{i1}")
wt_2[,2] <- paste0(wt_2[,2], "H_{i2}")
wt_2[,3] <- paste0(wt_2[,3], "H_{i3}")
lp_2 <- 
  cbind(
    format(coefs$model.2.bias, digits = 2),
    wt_2
  )
lp_2 <- apply(lp_2, 1, function(x) paste0(x, collapse = ""))
lp_2 <- paste0("Y_{i", 1:2, "}&=", lp_2, " \\notag")

cat("\\begin{align}")
cat(paste0(lp_2, collapse = " \\\\ \n"))
cat("\\end{align}")
```

Note that the coefficients’ signs are opposite for each of the outcome units. @fig-nnet-ex-layer-2 illustrates the transition from the hidden layer to the outcome layer. The weighted sum of the three variables at the top produces the regions of strong positive (or negative) activation for the two classes. Once again, the black curve shows where the zero contour is located. 

```{r}
#| label: fig-nnet-ex-layer-2
#| echo: false
#| out-width: 80%
#| fig-width: 7
#| fig-height: 5
#| warning: false
#| fig-cap: "Heatmaps of the three ReLU units and their output units across the predictor space."

output_layer <- 
  components |>
  select(-starts_with("lp"), -starts_with("relu_")) |>
  pivot_longer(cols = c(starts_with("output"))) |>
  ggplot(aes(A, B)) +
  geom_tile(aes(fill = value)) +
  geom_contour(
    # data = grid_pred,
    aes(z = value),
    breaks = 0,
    col = "black"
  ) +
  facet_wrap(~ name, nrow = 1) + 
  scale_fill_gradient2(high = "#420F75FF", low = "#EAC541FF") + 
  coord_fixed() + 
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  )  + 
  labs(title = "Output", y = NULL) 

relu_layer / output_layer
```

The two panels show nearly opposite patterns, which is not coincidental. The model converges to this structure so that the output units would favor one class or the other. The legend on the bottom right is clearly in the wrong units; we’d like them to resemble probabilities. As previously mentioned, the softmax function coerces these values to be on [0, 1] and to also sum to one. This produces the class boundary seen in @fig-nnet-ex. 

As previously stated, ReLU is not a smooth function, so this demonstration will be somewhat different from the other activation functions; they generally don’t completely isolate regions of the predictor space. However, the general flow of the network is still the same. 

### Training and Optimization {#sec-nnet-training}

It can be challenging to train neural network classifiers. Their model structure is composed of multiple nested nonlinear functions that are not smooth, continuous functions for activations such as ReLU. The loss function, commonly cross-entropy, is unlikely to be convex, especially as model complexity increases. As a consequence, we have no guarantees that a gradient-based optimizer will converge. There is also the possibility of local minima that might trap the training process, saddle points that could lead the search in the wrong direction, and regions of vanishing gradients. These issues will be discussed below. 

The complexity of these highly nested nonlinear regression models can create significant numerical issues. First is the challenge of local minima. In some cases, ML models can provide some guarantees regarding the existence of a global minimum loss value, generalized linear models being a good example. This may not be the case for neural networks. Their loss surface may contain multiple valleys where the derivatives of some of the parameters are zero, but a better solution exists in the parameter space. In other words, the model can appear to converge but has suboptimal performance. @baldi1989neural shows, under some conditions, that single-layer feedforward networks can have a unique global minimum. It should be expected that as more layers and units are added, the probability of a unique best solution decreases. Our optimization algorithm, when "caught" in a local valley, should be able to escape. 

A second related issue is saddle points. These are locations where all of the gradients are zero (i.e., a stationary point), but some directions have increasing loss, and others will produce a decrease. @dauphin2014identifying postulate that these will exist, and their propensity to occur increases with the number of model parameters. The worry here is that, depending on the type of multidimensional curvature that we face, the optimization algorithm might move in the direction of _increasing_ loss. 

@fig-stationary has an example with two parameters and a highly nonlinear loss surface. There are four stationary points, the best of which minimizes the loss function at $\boldsymbol{\theta} = [`r round(stationary$x[stationary$type == "minimum"], 1)`, `r round(stationary$y[stationary$type == "minimum"], 1)`]$. The lower right region has a point associated with maximum loss. The other two locations are saddle points. The point at $\boldsymbol{\theta} = [`r round(stationary$x[stationary$type == "saddle point" & stationary$x > 6], 1)`, `r round(stationary$y[stationary$type == "saddle point" & stationary$x > 6], 1)`]$ is a good example; moving to the left or right of the point will increase the loss. Moving up or down will generate a decrease in the loss (with the upper value being the correct choice in this parameter range). Our optimizer should be able to move in the correct direction or, like simulated annealing, be able to recover and move back to progressively better results. 



::: {#fig-stationary}

::: {.figure-content}

```{r}
#| label: stationary
#| echo: false
#| out-width: 70%
#| fig-width: 6
#| fig-height: 5

loss_grid |>
	rename(loss = res) |> 
	ggplot(aes(x, y)) +
	geom_tile(aes(fill = loss)) +
	geom_contour(aes(z = loss), col = "darkred", bins = 7, alpha = 3 / 4) +
	scale_fill_gradient(low = "white", high = "red") +
	scale_shape_manual(values = c(2, 6, 1)) +
	geom_point(
		data = stationary,
		aes(pch = type),
		cex = 2.5,
		col = "black"
	) + 
	coord_fixed(ratio = 5) + 
	labs(x = expression(theta[1]), y = expression(theta[2]))
```

:::

The loss surface for $\psi(\boldsymbol{\theta}) = \theta_1 cos(0.5\:\theta_1) - 2 \exp(-6(\theta_2 - 0.3)^2) + 0.2\: \theta_1 \theta_2$. Four stationary points are shown.

:::

Another possible factor is that many deep learning models are trained on vast amounts of data. This in itself prevents constraints on how much data can be held in memory at once, and other logistical issues. This may not be an issue for the average application that uses tabular data, but the problem still exists. 

Before proceeding, recall @sec-gradient-opt where gradient descent was introduced. There are two commonly used classes of gradient-based optimization methods: 

 - _First-order techniques_: $\quad\boldsymbol{\theta}_{i+1} = \boldsymbol{\theta}_i - \alpha\:g(\boldsymbol{\theta}_i)$,

- _Second-order methods_: $\quad\boldsymbol{\theta}_{i+1} = \boldsymbol{\theta}_i - \alpha H(\boldsymbol{\theta_i})^{-1}\:g(\boldsymbol{\theta}_i)$,

where $\boldsymbol{\theta}$ are the unknown parameters, $\alpha$ is the learning rate, $g(\boldsymbol{\theta}_i)$ is the gradient vector of first derivatives, and $H(\boldsymbol{\theta_i})$ is the Hessian matrix of second derivatives. The Hessian measures the curvature of the loss function at the current iteration. Pre-multiplying by its inverse rescales the gradient vector, generally making it more effective. We'll discuss them both in @sec-gradient-descent and @sec-newtons-method below. In most cases, second-order methods are preferred since they are more accurate. However, first-order search is more commonly used because it is faster to compute and can be much more memory efficient^[Making it infeasible to compute efficiently when the training set is very large.]

To demonstrate, an MLP model was used to model the forestation data that contained one layer of 30 ReLU units and an additional layer of 5 ReLU units, yeilding `r num_optimizer_param` model parameters. An L<sub>2</sub> penalty of 0.15 was used to regularize the cross-entropy loss function. A maximum of 100 epochs was specified, but training would stop after five bad epochs of validation loss. An initial learning rate of $\alpha = 0.1$ was used, but a variable, step-wise rate schedule was used to modulate the rate across epochs (described below). 

The model was first trained using gradient descent. For illustration, training was executed five times using different initial parameter values. The leftmost panel in @fig-optimizers shows the results. In each case, the loss function had an initial middling reduction. In each case, the model used all 100 epochs. This approach was not very successful. There was not much of a reduction in the loss function, and the terminal phase had a minuscule decrease with no real improvement. This _could_ be due to the search approaching the saddle point. In this case, the gradient becomes very close to zero, and first-order gradient searches have very small updates.

```{r}
#| label: fig-optimizers
#| echo: false
#| out-width: 90%
#| fig-width: 8
#| fig-height: 2.8
#| warning: false
#| fig-cap: "The validation loss function for the same MLP model for four optimzation strategies. The lines on each panel represents runs initialized with different random numbers. The profiles are normalized so that they start from the same initial loss value."

optimizers |> 
  ggplot(aes(epoch, loss_norm, col = seed, group = seed)) +
  geom_line(show.legend = FALSE,
            linewidth = 3 / 4,
            alpha = 1 / 2) +
  facet_wrap(~ method, scale = "free_x", nrow = 1) +
  scale_x_continuous(breaks = pretty_breaks()) + 
  labs(x = "Epochs", y = "Cross-Entropy (Validation)")
```

As a second approach, an approximation to Newton’s Method called the Limited-memory Broyden–Fletcher–Goldfarb–Shanno algorithm (L-BFGS) method. This uses an approximation to the Hessian matrix. In this case, there was an immediate and enormous reduction in loss in the first epoch. In most cases, the next few epochs resulted in further reductions. Some of the training runs quickly converged but one required 14 epochs. Compared to basic gradient descent, L-BFGS produced a much more effective model (based on the loss). The median time for optimization (per epoch) was `r round(optimizer_times$per_epoch[optimizer_times$method == "L-BFGS"], 3)`s, which was far slower than the median time for gradient descent (`r round(optimizer_times$per_epoch[optimizer_times$method == "GD"], 3)`s). However, the extra time is offset since L-BFGS is likely to use far fewer epochs in this scenario. 

As-is, gradient descent might be deemed ineffectual, and a second-order approach like L-BFGS might be the best approach. However, the plain version of gradient descent is almost never used. Instead, different variations of _stochastic_ gradient descent (SGD) are the norm. SGD uses randomly allocated _batches_ training set samples. The gradient search is updated by processing each of these batches until the entire trained set has been used, which signifies the end of that epoch. In other words, a potentially large number of parameter updates based on a small number of training point samples are used. The batch size used here was 64 data points. As a result, for this training set size, about 76 gradient updates are executed before the end of each epoch. 

As shown in @fig-optimizers, SGD does almost as well as L-BFGS, but most do not stop early; there is a relatively flat trend after about 20 epochs, where the search does not make substantial reductions of the loss. In terms of time per epoch, SGD was slightly better than L-BFGS at `r  round(optimizer_times$per_epoch[optimizer_times$method == "SGD"], 3)`s but will probably require more training. 

As we’ll see in the section below, many extensions to SGD have made it more effective. One improvement is to use _Nesterov momentum_ [@sutskever2013importance]. This initially computes a standard update to the parameters, but then approximates the gradient at this position to compute a second prospective step. The actual update that is used is the lookahead position. This technique shows results on par with L-BFGS with several repeats stopping early. However, the training time was longest, using `r round(optimizer_times$per_epoch[optimizer_times$method == "SGD with momentum"], 3)`s per epoch.  

::: {.note-box}
The main takeaways of this exercise are that training these models can be difficult and that the type of parameter optimization can substantially impact model quality. 
:::

Let’s discuss a few operational aspects of gradient-based methods before discussing the first- and second-order search techniques. 

#### Initialization {#sec-nnet-initialization}

Network parameter initialization can significantly impact training success. Given the size and complexity of these models, optimal methodologies for setting initial values are nearly impossible to derive. As such, parameters are initialized using random numbers. There are a few approaches to do this. @lecun2002efficient initialization uses a uniform distribution centered around zero with limits defined by the number of connections feeding into the current node (known as the "fan in" and is denoted as $p_{in}$). They used $U(\pm \sqrt{2/p_{in}})$ as the distribution. There have been a few variations of this: 

 - Glorot/Xavier initialization [@glorot2010understanding]: $U(\pm \sqrt{6/(p_{in}+p_{out})})$ 
 - He/Kaiming initialization [@he2015delving]: $N(0, \sqrt{2 / p_{in}})$.

One reason that initialization is important is to help mitigate numerical overflow errors. For example, the softmax transformation shown above exponentiates the raw output units (and sums them). If the output units have relatively large values, exponentiating and summing them can exceed allowable ranges. For example, the output units shown in @fig-nnet-ex-layer-2 have a wide enough range that can be as large as `r max(round(abs(components$output_1)), 0)`. 


#### Preprocessing {#sec-nnet-preprocessing}

Neural networks require complete, numeric features for computations. For non-numeric data, the tools described in @sec-categorical-predictors can be very effective. Additionally, there are pretrained neural network models that can translate the string to a word embedding [@Boykis_What_are_embeddings_2023], such as  BERT [@devlin2019bert] or GPT-4 [@achiam2023gpt]. The potential upside to doing this is that the resulting numeric features may measure some relative context between categorical values. For example, suppose we have a predictor with levels `red`, `darkred`, `green`, `chartreuse`, and `white`. Basic encoders, such as binary indicators for hashing features, would not recognize that there are three clusters in these values and that sets such as {`red`, `darkred`} and {`green`, `chartreuse`} should have values that are “near” one another. LLMs can measure this semantic similarity and potentially generate more meaningful features. 

If the non-numeric predictors are not complex, a layer could be added early in the network that enables the model to derive encodings based on the data. @guo2016entity describes a relatively simple architecture for doing this. While LLMs are overall better at creating embeddings, they are unaware of the _context_ in your data. Estimating the parameters associated with a simple embedding layer might be more effective since it is optimized for your data set. 

For numeric predictors, we suggest avoiding irrelevant predictors and highly collinear feature sets. @fig-irrelevant-predictors in @sec-feature-selection demonstrated the harm from including predictors with no relationship to the outcome into a neural network. Multicollinearity can also affect these models, even if we do not invert the feature matrix or use the Hessian during optimization. Highly redundant predictors can result in the loss function having difficult characteristics. Using an L<sub>2</sub> penalty or a feature extraction method can be very helpful. 

Also, predictors with highly skewed distributions might benefit from a transformation to make their distribution more symmetric. If not, outlying values can exhibit significant leverage in computations, which can prevent high-quality parameter estimates. 

Finally, since the networks use random initialization, we should ensure that our predictors are standardized to have the same units. If there are only numeric predictors, centering/scaling or orderNorm operations can be very effective^[orderNorm is perhaps more attractive since it standardizes the predictors and coerces them to a symmetric distribution.]. If we use this approach, we must also apply these same standardizations to binary indicators that are created from categorical predictors.  Alternatively, we could leave 0/1 indicators as-is but _range_ scale our predictors to also be between zero and one. This may have a disadvantage; the range will be common acorss predictors but the mean and variance can be quite different. 


#### Learning Rates {#sec-nnet-learning-rates}

The learning rate ($\alpha$) can have a profound effect on model quality, perhaps more than the structural tuning parameters that define complexity (e.g., number of hidden units). A large learning range will help converge towards effective parameter estimates, but once there, it can cause the model to consistently overshoot the best. One approach to modulating the rate is to use a scheduler [@lu2022gradient,chapt 4]. These are functions/algorithms that will change the rate (often to decrease it) over epochs. For example, to monotonically decrease the rate, two options are:

- inverse decay: $f(\alpha, \delta, i) = \alpha_0/(1 + \delta i)$ 
- exponential decay: $f(\alpha, \delta, i) = \alpha_0\exp(-\delta i)$

where $i$ is the epoch number and $\delta$ is a decay value. These two functions gradually determine the rate so that, when near an optimum, it is slow enough not to exceed the optimal settings. Another approach is a step function that decreases the rate but holds it constant for a range of epochs. Those ranges can be predefined or can be set dynamically so that the rate decreases once the loss function is reduced. The function for the former is: 

- step decay: $f(\alpha, \zeta, i) = \alpha_0 \zeta^{\lfloor i/r\rfloor}$ 

where $\zeta$ is the reduction value, $r$ is the step length, and $\lfloor x\rfloor$ is the floor function. 

Finally, there are schedulers that modulate the rate up and down. A cyclic schedule [@smith2017cyclical] initially starts at a maximum value, decreases to a minimum value, and then cyclically increases again. For some range $r$ and minimum rate $\alpha_0$, an epoch's location within the series is $cycle = \lfloor 1 + (i / (2r) )\rfloor$ and the rate is

- cyclic: $f(\alpha_0, \alpha, r) = \alpha_0 + ( \alpha - \alpha_0 ) \max(0, 1 - x)$

where $x = |(i/r) - (2\:cycle) + 1|$. This is a good option when we are not sure how fast or slow the convergence will be. If the rate is too high near an optimum value, it will soon decrease to an appropriate value. Also, if the search is stuck in a local optimum (or a saddle-point), moving back to a larger learning rate might help it escape. 

@fig-schedulers uses the loss function example from @fig-stationary to demonstrate the effect of rate schedulers. Two starting points are used. One (in orange) is very close to a saddle point, while the other starts from the upper right corner of the parameter space (in green). 

```{r}
#| label: fig-schedulers
#| echo: false
#| out-width: 70%
#| fig-width: 5.5
#| fig-height: 5.5
#| fig-cap: "The effect of different schedulers on the rate and success of convergence."
base_sgd_plot(rates, NULL) + facet_wrap(~schedule)
```

Using a constant learning rate of $\alpha = 0.1$, the search marches to the optimum value, but oscillates wildly on approach. In the end, it does not come close to the best estimates since it is constrained to make large jumps on each iteration. The decay panel represents inverse decay with $\delta = 0.025$. For the corner starting point, this is a very effective strategy. From the saddle point, we can see that the decrease in the learning rate results in fewer oscillations as the search proceeds and eventually finds its way to the optimum. The step panel used an initial learning rate of $\alpha_0 = 0.1$, $\eta = 0.5$, and a step length of 20 epochs. This worked well when the search began from the saddle location, but it did poorly when starting at the corner. The success depended on when the search was near the optimum; if not timed right, the oscillations were too wide to converge on the best point. A reduction in the step length might help. Finally, the cyclic schedule allowed rates to move between $\alpha_0 = 0.001$ to $\alpha = 0.1$ with a cycle range of $r = 10$ epochs. This appears to work well or, at least, had good timing. 

The type of schedule, and its additional parameter(s), are added to the list of possible values that could be optimized during tuning.


#### Early Stopping {#sec-early-stopping}

#### Batch Learning {#sec-batch-learning}

#### Gradient Descent {#sec-gradient-descent}

#### Newton’s Method {#sec-newtons-method}

### Forestation Modeling  {#sec-nnet-forestation}

## Special Tabular Network Models  {#sec-tab-net}

### Transformers {#sec-nnet-transformers}

### Attention {#sec-nnet-attention}

## K-Nearest Neighbors  {#sec-cls-knn}

### Discriminant Adaptive Nearest Neighbors  {#sec-dann}

## Support Vector Machines {#sec-cls-svm}

## Chapter References {.unnumbered}
